---
title: "Workbook — snRNA-seq Preprocessing"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

# Workbook — Preprocessing the EO snRNA-seq Data

This is your hands-on working surface for Friday. Run chunks one at a time. Where you see `___`, fill in the value — the chunk will fail informatively if you leave a blank.

When you get stuck, the full episode at the same name on the course website is the answer key. Don't peek before you predict.

> **Setup check before you start:** Confirm `getwd()` returns `~/ssrnaseq` or similar; you should see `data/ssrnaseq_data/EO/` in the file pane. If not, set the working directory via Session → Set Working Directory → To Project Directory.

## Setup

```{r load-libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(glmGamPoi)
library(presto)
library(harmony)
```

## Load the four EO samples

```{r load-data, message=FALSE, warning=FALSE}
tissue <- "EO"
sample_names <- c("BB48", "BB49INJ", "BB50", "BB50INJ")

list_seurat <- list()

for (sample in sample_names) {
  data_dir <- file.path("data/ssrnaseq_data", tissue, sample)

  seurat_data <- Read10X(data.dir = data_dir)
  seurat_obj <- CreateSeuratObject(
    counts = seurat_data,
    min.features = 1,
    project = paste0(tissue, "_", sample)
  )

  list_seurat[[sample]] <- seurat_obj
}

# Sanity check: print the number of nuclei per sample
sapply(list_seurat, ncol)
```

## Merge samples and stamp metadata

```{r merge-and-annotate, message=FALSE, warning=FALSE}
merged_seurat <- merge(
  x = list_seurat[[1]],
  y = list_seurat[2:length(list_seurat)],
  add.cell.id = names(list_seurat)
)
merged_seurat <- JoinLayers(merged_seurat)

merged_seurat$sample <- sub("EO_", "", merged_seurat$orig.ident)
merged_seurat$fish <- merged_seurat$sample
```

> **Predict first.** Below, we map fish to treatment. From memory (or the implant-11kt episode): which two fish received 11-KT, and which two received vehicle? Write your prediction in the comment line before filling in the gaps.
>
> _Your prediction:_

```{r fill-treatment-map, eval=FALSE}
# Your prediction: BB__ and BB__ are 11-KT; BB__ and BB__ are vehicle
treatment_map <- c(
  BB48    = ___,
  BB49INJ = ___,
  BB50    = ___,
  BB50INJ = ___
)
merged_seurat$treatment <- unname(treatment_map[merged_seurat$sample])
head(merged_seurat@meta.data)
```

::: {.callout-important title="Challenge — The INJ suffix"}

Two samples have an `INJ` suffix in their names. What does it mean? Why is it dangerous to use the suffix to infer treatment?

::: {.callout-tip title="Hint / solution" collapse="true"}
`INJ` is a library-preparation label, not a treatment label. BB49INJ is 11-KT; BB50INJ is vehicle. Inferring treatment from the string would flip BB49INJ's label — opposite of the truth. Always use fish identity, never sample name strings.
:::

:::

## Visual QC check

```{r qc-violin, fig.width=10, fig.height=5, message=FALSE, warning=FALSE}
p_count <- VlnPlot(merged_seurat,
  features = "nCount_RNA",
  group.by = "orig.ident",
  pt.size = 0
) + theme(legend.position = "none") + ggtitle("UMI counts per nucleus")

p_feat <- VlnPlot(merged_seurat,
  features = "nFeature_RNA",
  group.by = "orig.ident",
  pt.size = 0
) + theme(legend.position = "none") + ggtitle("Genes detected per nucleus")

p_count | p_feat
```

> **Predict first.** Do any samples look categorically different from the others (as Pool 4 did from Pools 1–2 in the QC video)?
>
> _Your prediction:_

## SCTransform normalization

```{r sctransform, message=FALSE, warning=FALSE}
merged_seurat <- SCTransform(merged_seurat, verbose = FALSE)
```

## PCA and the elbow plot

```{r run-pca, message=FALSE, warning=FALSE}
merged_seurat <- RunPCA(merged_seurat, verbose = FALSE, assay = "SCT")
```

```{r elbow-plot, message=FALSE, warning=FALSE}
ElbowPlot(merged_seurat, ndims = 30)
```

> **Predict first.** Look at the elbow plot. Where do you think the elbow falls — PC 10? PC 20? PC 30? Write your guess in plain language before running the next chunk.
>
> _Your prediction:_ _____ PCs

```{r set-n-pcs, eval=FALSE}
n_pcs <- ___   # your choice based on the elbow plot
```

::: {.callout-important title="Challenge — Why does the elbow exist?"}

What does each principal component represent? Why does the curve flatten ("the elbow") at some point?

::: {.callout-tip title="Hint / solution" collapse="true"}
Each PC is a weighted combination of genes that captures one axis of variance. Early PCs capture the dominant biological axes (cell-type differences). Later PCs capture progressively smaller, noisier components. Past the elbow, additional PCs are mostly noise; including them adds little signal but a lot of computational cost.
:::

:::

```{r run-umap-pca, message=FALSE, warning=FALSE}
# Uses n_pcs from the previous chunk; defaults to 20 if you skipped the gap
if (!exists("n_pcs")) n_pcs <- 20

merged_seurat <- RunUMAP(
  merged_seurat,
  dims           = 1:n_pcs,
  reduction      = "pca",
  reduction.name = "umap.pca",
  verbose        = FALSE
)
```

```{r save-checkpoint-01, message=FALSE, warning=FALSE}
# Save the post-SCT/PCA state in case you need to back out of later changes
dir.create("data/checkpoints", showWarnings = FALSE, recursive = TRUE)
saveRDS(merged_seurat, file = "data/checkpoints/01_merged_sct.rds")
```

## Harmony integration

```{r harmony-integration, message=FALSE, warning=FALSE}
merged_seurat <- RunHarmony(
  merged_seurat,
  group.by.vars  = "orig.ident",
  reduction      = "pca",
  reduction.save = "harmony",
  verbose        = FALSE
)
```

## Cluster on Harmony

> **Predict first.** Resolution sets cluster granularity. Lower = fewer/broader clusters; higher = more/finer. For an EO dataset where you expect 1 dominant electrocyte type plus 3–5 smaller types (muscle, Schwann, endothelial, motor neurons, maybe stalk vs. face), what resolution would you start with?
>
> _Your prediction:_ resolution = _____

```{r fill-resolution, eval=FALSE}
my_resolution <- ___   # your choice

merged_seurat <- FindNeighbors(merged_seurat,
  dims      = 1:n_pcs,
  reduction = "harmony",
  verbose   = FALSE
)
merged_seurat <- FindClusters(merged_seurat,
  resolution = my_resolution,
  verbose    = FALSE
)

merged_seurat <- RunUMAP(
  merged_seurat,
  dims           = 1:n_pcs,
  reduction      = "harmony",
  reduction.name = "umap.harmony",
  verbose        = FALSE
)
```

## Before / after Harmony

```{r harmony-comparison, fig.width=14, fig.height=8, message=FALSE, warning=FALSE}
p1 <- DimPlot(merged_seurat, reduction = "umap.pca",
  group.by = "orig.ident", alpha = 0.4) + ggtitle("PCA — by sample")
p2 <- DimPlot(merged_seurat, reduction = "umap.pca",
  group.by = "treatment", alpha = 0.4,
  cols = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) + ggtitle("PCA — by treatment")
p3 <- DimPlot(merged_seurat, reduction = "umap.harmony",
  group.by = "orig.ident", alpha = 0.4) + ggtitle("Harmony — by sample")
p4 <- DimPlot(merged_seurat, reduction = "umap.harmony",
  group.by = "treatment", alpha = 0.4,
  cols = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) + ggtitle("Harmony — by treatment")

(p1 | p2) / (p3 | p4)
```

::: {.callout-important title="Challenge — Did Harmony help?"}

Compare the two rows of UMAPs.

1. In the "by sample" panels, do sample-specific islands present in the PCA UMAP dissolve in the Harmony UMAP? Or were they already mixed?
2. In the "by treatment" panels, do 11-KT and vehicle stay partially separated after Harmony? They should.
3. Based on what you see, did Harmony provide assurance (small change) or active correction (big change)?

::: {.callout-tip title="Hint / solution" collapse="true"}
For this low-batch dataset, the PCA and Harmony UMAPs often look very similar. That's a good result — minimal technical batch is exactly what we want. Treatment separation should be preserved across both. The full episode walks through what each panel tells you.
:::

:::

## Resolution sweep

::: {.callout-important title="Challenge — Pick a resolution"}

Re-run clustering at three resolutions and visualize. Then set it back to your chosen value.

```{r resolution-sweep, eval=FALSE}
for (res in c(0.3, 0.8, 1.5)) {
  merged_seurat <- FindClusters(merged_seurat, resolution = res, verbose = FALSE)
  print(
    DimPlot(merged_seurat, reduction = "umap.harmony",
            group.by = "seurat_clusters", label = TRUE) +
      ggtitle(paste("resolution =", res))
  )
}

# After exploring, restore your chosen resolution so Saturday is reproducible
merged_seurat <- FindClusters(merged_seurat, resolution = ___, verbose = FALSE)
```

Which resolution gives groups that you'd be willing to start annotating tomorrow?

::: {.callout-tip title="Hint / solution" collapse="true"}
The full episode uses `resolution = 0.8`. Lower resolutions merge electrocyte sub-types; higher resolutions over-split them. ~10–14 clusters at 0.8 is a defensible starting point. The "right" value depends on the biological question.
:::

:::

## Three-panel final view

```{r umap-cluster-views, fig.width=15, fig.height=5, message=FALSE, warning=FALSE}
p_clu <- DimPlot(merged_seurat, reduction = "umap.harmony",
  group.by = "seurat_clusters", label = TRUE, repel = TRUE) +
  ggtitle("Cluster") + theme(legend.position = "none")
p_smp <- DimPlot(merged_seurat, reduction = "umap.harmony",
  group.by = "orig.ident", alpha = 0.4) + ggtitle("Sample")
p_trt <- DimPlot(merged_seurat, reduction = "umap.harmony",
  group.by = "treatment", alpha = 0.4,
  cols = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) + ggtitle("Condition")

p_clu | p_smp | p_trt
```

## Save the checkpoint for Saturday

```{r save-checkpoint-02, message=FALSE, warning=FALSE}
saveRDS(merged_seurat, file = "data/checkpoints/02_harmony_clustered.rds")
cat("Friday checkpoint saved.\n")
```

> **Done.** Tomorrow's workbook (`workbook/snrnaseq-annotation-and-de.Rmd`) starts by loading this file. If you arrive Saturday and find you missed today's session, run the first chunk of the Saturday workbook — it loads the pre-built checkpoint shipped on the server.
