---
title: "Workbook — Cell-Type Annotation & Differential Expression"
output: html_notebook
editor_options:
  chunk_output_type: inline
---

# Workbook — Saturday

Today's work is denser than yesterday's. The gaps land at decision points: which cell-type labels to trust, which paralog drives the stalk signal, which gene to defend as a candidate hit, and how to read the cross-study scatter plot.

The full episode at the same name on the course website is the answer key. Predict before you peek.

## Load Friday's checkpoint

```{r load-libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(glmGamPoi)
library(presto)
library(DESeq2)
library(limma)
library(ggrepel)
if (requireNamespace("speckle", quietly = TRUE)) library(speckle)
```

```{r load-checkpoint, message=FALSE, warning=FALSE}
# Loads the post-Harmony, post-clustering state you produced Friday.
# If you missed Friday, this loads the pre-built checkpoint shipped on the server.
merged_seurat <- readRDS("data/checkpoints/02_harmony_clustered.rds")
merged_seurat
```

## Find cluster markers

```{r prep-markers, message=FALSE, warning=FALSE}
merged_seurat <- PrepSCTFindMarkers(merged_seurat, verbose = FALSE)
```

```{r find-markers, message=FALSE, warning=FALSE}
all_markers <- FindAllMarkers(
  merged_seurat,
  only.pos = TRUE,
  verbose  = FALSE
)
all_markers <- all_markers |> arrange(cluster, desc(avg_log2FC))
head(all_markers, 20)
```

## Join NCBI gene-product descriptions

```{r parse-gtf, message=FALSE, warning=FALSE}
gtf_raw <- read_tsv(
  "data/ssrnaseq_data/genes.gtf.gz",
  comment        = "#",
  col_names      = FALSE,
  col_types      = "ccciicccc",
  show_col_types = FALSE
)
gene_desc <- gtf_raw |>
  filter(str_detect(X9, 'product "')) |>
  mutate(
    gene_id   = str_extract(X9, 'gene_id "([^"]+)"', group = 1),
    gene_name = str_extract(X9, 'gene "([^"]+)"', group = 1),
    product   = str_extract(X9, 'product "([^"]+)"', group = 1)
  ) |>
  dplyr::select(gene_id, gene_name, product) |>
  distinct(gene_id, .keep_all = TRUE)

gene_lookup <- bind_rows(
  gene_desc |> transmute(key = gene_id, gene_name, product),
  gene_desc |>
    filter(!is.na(gene_name), gene_name != gene_id) |>
    transmute(key = gene_name, gene_name, product)
) |>
  distinct(key, .keep_all = TRUE)

annotated_markers <- all_markers |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key")
  )

annotated_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 3) |>
  ungroup() |>
  dplyr::select(cluster, gene, gene_name, product, avg_log2FC, p_val_adj)
```

> **Predict first.** Scan the top markers. Before running module scoring, which cluster do you think will be electrocyte? Which is most likely Schwann? Write your guesses, including the cluster numbers.
>
> _Your prediction:_

## Curated marker gene lists

The marker lists below come from the published EO literature (Gallant et al. 2014, Thompson et al. 2018, Losilla & Gallant 2025, Baumann et al. 2025, others). They are intentionally not gapped — the curation is the literature's contribution, not yours.

```{r marker-gene-lists, message=FALSE, warning=FALSE}
marker_genes <- list(
  Electrocyte = c(
    "LOC125742960", "scn4ab",       # Nav1.4 paralogs
    "LOC125743019", "kcna7", "kcnq1.2", "kcnq5a", "kcnn2",   # K+ channels
    "cnga3a", "atp1b2b",
    "rrad", "igsf9ba", "robo2", "tenm1"
  ),
  Muscle = c("pax7b", "pitx3", "myom2a", "myom2b", "hey2", "ezh2",
             "megf10", "mcama", "fgfr4", "tbx22"),
  Schwann = c("col28a1a", "smoc1", "prss12", "antxr1b",
              "numb", "olfml2ba", "col6a3", "sema3bl"),
  Endothelial = c("cdh5", "pecam1b", "kdr", "flt1",
                  "clec14a", "pcdh12", "podxl", "adgrl4"),
  Motorneuron = c("mnx1", "LOC125712540", "LOC125745880",
                  "LOC125704285", "LOC125706047",
                  "chata", "slc18a3a", "elavl3", "rbfox3a",
                  "nefma", "nefla", "neflb", "LOC125751823", "prph",
                  "sv2a", "snap25a", "LOC125712969",
                  "gap43", "nrn1b", "nrxn1a")
)
marker_genes <- lapply(marker_genes, function(g) g[g %in% rownames(merged_seurat)])
print(sapply(marker_genes, length))

pmca_genes <- c("atp2b1a", "atp2b2", "atp2b3b", "atp2b4")
pmca_genes <- pmca_genes[pmca_genes %in% rownames(merged_seurat)]
```

## Module scores per cell type

```{r module-score, fig.width=12, fig.height=9, message=FALSE, warning=FALSE}
for (ct in names(marker_genes)) {
  if (length(marker_genes[[ct]]) < 3) next
  merged_seurat <- AddModuleScore(
    merged_seurat,
    features = list(marker_genes[[ct]]),
    name     = paste0(ct, "_score"),
    nbin     = 12, seed = 42
  )
  old_col <- paste0(ct, "_score1")
  new_col <- paste0(ct, "_score")
  merged_seurat@meta.data[[new_col]] <- merged_seurat@meta.data[[old_col]]
  merged_seurat@meta.data[[old_col]] <- NULL
}
score_cols <- paste0(names(marker_genes), "_score")
score_cols <- score_cols[score_cols %in% colnames(merged_seurat@meta.data)]

if (length(pmca_genes) >= 2) {
  merged_seurat <- AddModuleScore(
    merged_seurat, features = list(pmca_genes),
    name = "PMCA_score", nbin = 12, seed = 42
  )
  merged_seurat$PMCA_score <- merged_seurat$PMCA_score1
  merged_seurat$PMCA_score1 <- NULL
}

FeaturePlot(merged_seurat,
  features  = c(score_cols, "PMCA_score"),
  reduction = "umap.harmony",
  ncol = 2, order = TRUE
)
```

::: {.callout-important title="Challenge — Compare your prediction"}

Look at the module-score Feature plots. Did your predicted electrocyte cluster light up brightest on `Electrocyte_score`? Did your predicted Schwann cluster light up on `Schwann_score`? Where were you wrong, and what tipped you off in the top markers?

::: {.callout-tip title="Hint / solution" collapse="true"}
Electrocyte signatures are usually strong and central on the UMAP because electrocytes dominate the tissue. The PMCA score lights up a smaller subset within the electrocyte territory — that's your stalk candidate. Schwann, endothelial, and motor-neuron territories are usually smaller and peripheral on the UMAP.
:::

:::

## Assign cluster labels and refine stalk identity

```{r annotate-from-markers, fig.width=9, fig.height=6, message=FALSE, warning=FALSE}
cluster_scores <- merged_seurat@meta.data |>
  group_by(seurat_clusters) |>
  summarise(across(all_of(score_cols), mean, .names = "{.col}"), .groups = "drop")

cluster_labels <- cluster_scores |>
  pivot_longer(-seurat_clusters, names_to = "score_col", values_to = "mean_score") |>
  mutate(cell_type = str_remove(score_col, "_score")) |>
  group_by(seurat_clusters) |>
  slice_max(mean_score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  dplyr::select(seurat_clusters, cell_type, mean_score) |>
  arrange(as.integer(as.character(seurat_clusters)))

label_vec <- setNames(cluster_labels$cell_type,
  as.character(cluster_labels$seurat_clusters))
merged_seurat$cell_type_go <- unname(label_vec[as.character(merged_seurat$seurat_clusters)])

# Stalk refinement
merged_seurat$cell_type_detail <- merged_seurat$cell_type_go
if (length(pmca_genes) >= 2 && "PMCA_score" %in% colnames(merged_seurat@meta.data)) {
  stalk_scores <- merged_seurat@meta.data |>
    group_by(seurat_clusters) |>
    summarise(
      mean_PMCA = mean(PMCA_score),
      mean_Elec = mean(Electrocyte_score),
      n_cells = n(),
      initial_label = names(sort(table(cell_type_go), decreasing = TRUE))[1],
      .groups = "drop"
    ) |>
    mutate(stalk_score = mean_PMCA * pmax(mean_Elec, 0)) |>
    arrange(desc(stalk_score))
  stalk_cluster <- as.character(stalk_scores$seurat_clusters[1])

  pmca_by_cluster <- merged_seurat@meta.data |>
    filter(cell_type_go == "Electrocyte") |>
    group_by(seurat_clusters) |>
    summarise(mean_PMCA = mean(PMCA_score), n_cells = n(), .groups = "drop") |>
    arrange(desc(mean_PMCA))

  stalk_idx <- rownames(merged_seurat@meta.data)[
    as.character(merged_seurat$seurat_clusters) == stalk_cluster
  ]
  merged_seurat$cell_type_detail[stalk_idx] <- "Stalk_Electrocyte"
}

DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "cell_type_detail",
  label     = TRUE, repel = TRUE
) + ggtitle("EO cell types") + NoLegend()
```

```{r save-checkpoint-03, message=FALSE, warning=FALSE}
saveRDS(merged_seurat, file = "data/checkpoints/03_annotated.rds")
```

## Identify the stalk-driver PMCA paralog

> **Predict first.** Four PMCA paralogs (`atp2b1a`, `atp2b2`, `atp2b3b`, `atp2b4`) are present. The module score is the **average** of all four. Which single paralog do you think will turn out to be the most stalk-specific? You'll find out by looking at the FeaturePlot below.
>
> _Your prediction:_

```{r pmca-paralogs, fig.width=12, fig.height=5, message=FALSE, warning=FALSE}
FeaturePlot(merged_seurat,
  features = pmca_genes,
  reduction = "umap.harmony",
  ncol = 2, order = TRUE
)
```

::: {.callout-important title="Challenge — Single-paralog stalk marker"}

Looking at the four PMCA FeaturePlots:

1. Which paralog has the most stalk-restricted expression (lights up only in the stalk cluster)?
2. Which paralog is most broadly expressed (lights up everywhere)?
3. If you had to recommend a **single** PMCA paralog as a stalk marker for a future in-situ hybridization study, which one and why?

::: {.callout-tip title="Hint / solution" collapse="true"}
The Baumann 2025 paper used a pan-PMCA antibody, so the specific paralog is open. The single-paralog story is precisely what snRNA-seq can resolve. You're looking for tight, stalk-specific signal — the paralog whose territory looks most like the Stalk_Electrocyte label.
:::

:::

## Pseudobulk DE

> **Predict first.** Before we aggregate, write down the grouping variable for `AggregateExpression`. We want one count per gene per fish per cell type. What two metadata columns does `group.by` need?
>
> _Your prediction:_ `group.by = c(___, ___)`

```{r aggregate-expression, eval=FALSE}
pseudobulk_counts <- AggregateExpression(
  merged_seurat,
  assays        = "RNA",
  group.by      = c(___, ___),
  return.seurat = FALSE
)$RNA

cat("Pseudobulk matrix:", dim(pseudobulk_counts), "\n")
print(colnames(pseudobulk_counts))
```

::: {.callout-important title="Challenge — Why these two columns?"}

Defend your choice for the two grouping variables. What goes wrong if you use only one? What goes wrong if you use a third?

::: {.callout-tip title="Hint / solution" collapse="true"}
We need `(cell_type × sample)` — one count per gene per cell type per fish. With only `cell_type`, you'd pool all four fish into one number per cell type, losing the n=2-per-group structure DESeq2 needs. With only `sample`, you'd pool all cell types within each fish into one number, losing cell-type specificity. Adding a third grouping variable (e.g., `treatment`) is redundant — treatment is already a property of the sample.

The full episode uses `c("cell_type_go", "orig.ident")`.
:::

:::

## Build the metadata and run DESeq2

```{r pseudobulk-pipeline, message=FALSE, warning=FALSE, eval=FALSE}
# Use the value you set above
if (!exists("pseudobulk_counts")) {
  pseudobulk_counts <- AggregateExpression(
    merged_seurat,
    assays        = "RNA",
    group.by      = c("cell_type_go", "orig.ident"),
    return.seurat = FALSE
  )$RNA
}

pb_meta <- data.frame(
  col_id = colnames(pseudobulk_counts),
  stringsAsFactors = FALSE
) |>
  tidyr::separate(col_id, into = c("cell_type", "orig_ident"),
                  sep = "_", extra = "merge") |>
  mutate(
    sample = sub("EO-", "", orig_ident),
    treatment = case_when(
      sample %in% c("BB48", "BB49INJ") ~ "11kt",
      TRUE ~ "vehicle"
    ),
    treatment = factor(treatment, levels = c("vehicle", "11kt"))
  )
rownames(pb_meta) <- colnames(pseudobulk_counts)

sample_counts_per_ct <- pb_meta |>
  group_by(cell_type, treatment) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(cell_type) |>
  filter(all(n >= 2)) |>
  pull(cell_type) |>
  unique()

de_results <- list()
dds_objects <- list()

for (ct in sample_counts_per_ct) {
  cols <- which(pb_meta$cell_type == ct)
  mat <- pseudobulk_counts[, cols, drop = FALSE]
  meta_sub <- pb_meta[cols, , drop = FALSE]
  mat <- mat[rowSums(mat) > 0, , drop = FALSE]

  dds <- DESeqDataSetFromMatrix(
    countData = mat, colData = meta_sub, design = ~treatment
  )
  suppressWarnings(dds <- DESeq(dds, quiet = TRUE))
  res <- results(dds,
    contrast = c("treatment", "11kt", "vehicle"), alpha = 0.05)

  de_results[[ct]] <- as.data.frame(res) |>
    tibble::rownames_to_column("gene") |>
    mutate(cell_type = ct) |>
    arrange(padj)
  dds_objects[[ct]] <- dds
}

de_combined <- bind_rows(de_results)
saveRDS(list(de_results = de_results, dds_objects = dds_objects, pb_meta = pb_meta),
  file = "data/checkpoints/04_pseudobulk_dds.rds")
```

## Top DE table and per-fish visualization

```{r de-table, message=FALSE, warning=FALSE}
de_combined |>
  filter(!is.na(padj), padj < 0.1) |>
  group_by(cell_type) |>
  slice_min(padj, n = 5) |>
  ungroup() |>
  left_join(gene_lookup |> dplyr::select(key, product), by = c("gene" = "key")) |>
  dplyr::select(cell_type, gene, product, log2FoldChange, lfcSE, padj)
```

```{r pseudobulk-visualization, fig.width=11, fig.height=5, message=FALSE, warning=FALSE}
ct_vis <- if ("Electrocyte" %in% sample_counts_per_ct) "Electrocyte" else sample_counts_per_ct[1]
dds_vis <- dds_objects[[ct_vis]]
dds_vis <- estimateSizeFactors(dds_vis)
norm_mat <- counts(dds_vis, normalized = TRUE)

top_genes <- de_results[[ct_vis]] |>
  filter(!is.na(padj), padj < 1) |>
  slice_min(padj, n = 3) |>
  pull(gene)

norm_tidy <- norm_mat[top_genes, , drop = FALSE] |>
  as.data.frame() |>
  tibble::rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "col_id", values_to = "norm_count") |>
  left_join(pb_meta |> tibble::rownames_to_column("col_id"), by = "col_id") |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key")) |>
  mutate(label = if_else(is.na(product) | product == "", gene,
    paste0(gene, "\n(", str_trunc(product, 40), ")")))

ggplot(norm_tidy, aes(x = treatment, y = norm_count, color = treatment)) +
  geom_point(size = 4, position = position_jitter(width = 0.08, seed = 42)) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  scale_color_manual(values = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) +
  theme_bw() +
  theme(legend.position = "none", strip.text = element_text(size = 8)) +
  labs(title = paste("Pseudobulk —", ct_vis), y = "DESeq2-normalized counts", x = "Treatment")
```

::: {.callout-important title="Challenge — Defend one gene"}

Pick **one** top DE gene from the plot above and answer:

1. Are both 11-KT fish higher (or both lower) than both vehicle fish?
2. Look up the `product` description — does it make biological sense for an 11-KT response?
3. What would you do next to validate it?

Write your defense in the space below.

> _Your candidate gene:_
>
> _Replicate consistency:_
>
> _Biological plausibility:_
>
> _Next validation step:_

::: {.callout-tip title="Hint / solution" collapse="true"}
The honest version: only call a gene a "candidate" if both replicates in each condition agree in direction. The episode discusses ion channels (Na⁺, K⁺, Ca²⁺), sarcomeric proteins, and systemic androgen responses (lipid metabolism). Validation next steps: qPCR on independent fish; in-situ hybridization; comparison against published androgen-response datasets.
:::

:::

## Cross-study validation against Losilla & Gallant 2025

```{r load-losilla-data, message=FALSE, warning=FALSE}
losilla_bulk <- read_csv(
  "data/ssrnaseq_data/EO/losilla_et_al_data.csv",
  show_col_types = FALSE
) |>
  dplyr::rename(
    gene_bulk = "Gene Symbol",
    desc_bulk = "Gene Description",
    logFC_bulk = "logFC",
    FDR_bulk = "FDR"
  )

ct_check <- if ("Electrocyte" %in% names(de_results)) "Electrocyte" else names(de_results)[1]
overlap_tbl <- de_results[[ct_check]] |>
  inner_join(losilla_bulk, by = c("gene" = "gene_bulk")) |>
  mutate(concordant = sign(log2FoldChange) == sign(logFC_bulk)) |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key"))

n_detected <- nrow(overlap_tbl)
n_concordant <- sum(overlap_tbl$concordant, na.rm = TRUE)
cat(sprintf("Losilla DEGs detected: %d (concordance %.0f%%)\n",
  n_detected, 100 * n_concordant / n_detected))
```

> **Predict first.** Based on the printed concordance rate, what's your verdict — did our pipeline preserve the androgen signal? Write your prediction before reading the next callout.
>
> _Your prediction:_

```{r logfc-scatter, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
label_genes <- c("actn1", "actc1c", "myh6", "acta1b", "xirp1",
                 "elovl7a", "fhl1a", "atp1b2b",
                 "LOC125742960", "LOC125743019", "kcna7", "scn4b")

overlap_tbl |>
  mutate(label = if_else(gene %in% label_genes, gene, NA_character_)) |>
  ggplot(aes(x = logFC_bulk, y = log2FoldChange, color = concordant)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(aes(label = label), na.rm = TRUE,
    size = 3, max.overlaps = 20, box.padding = 0.4) +
  scale_color_manual(
    values = c("TRUE" = "#009E73", "FALSE" = "#CC4678"),
    labels = c("TRUE" = "Concordant", "FALSE" = "Discordant"),
    na.value = "grey60"
  ) +
  labs(x = "Bulk logFC (Losilla & Gallant 2025, 17αMT)",
       y = "Pseudobulk log2FC (this study, 11-KT)",
       color = "Direction") +
  theme_bw(base_size = 12)
```

::: {.callout-important title="Challenge — Interpret the scatter"}

1. Is concordance well above 50% (random chance)?
2. Do genes with the largest bulk logFC also have the largest snRNA-seq log2FC? What does that imply?
3. Losilla used 17αMT; we used 11-KT. List two biological reasons why some genes might respond differently across the two androgens.

::: {.callout-tip title="Hint / solution" collapse="true"}
The full episode discusses (a) receptor selectivity — 11-KT is the physiological teleost androgen, 17αMT is synthetic — and (b) aromatization — 17αMT can't be converted to estrogens whereas 11-KT can. Both could differentiate the gene response sets.
:::

:::

## Wrap

You've gone from raw counts to an annotated, integrated, hormone-contrasted EO atlas with cross-study validation. The most important habit to take away: **report direction and replicate consistency before padj** in any small-sample experiment.
