---
title: 'Examining EOD Duration from Raw Recordings'
teaching: 20
exercises: 25
---

::: {.callout-note title="Questions"}

- How do we load JSON EOD recordings into R?
- How is EOD duration measured from a raw waveform?
- Did 11-ketotestosterone treatment change EOD duration over time?

:::

::: {.callout-tip title="Objectives"}

- Load JSON recording files into R using the `jsonlite` package
- Navigate a list of recording objects to extract waveform data and metadata
- Visualize a raw EOD waveform using `ggplot2`
- Measure EOD duration by applying a threshold to a normalized waveform
- Build a tidy dataset by applying a duration function across many files
- Interpret a faceted comparison plot of EOD duration across treatment groups

:::

## Introduction

In the previous episodes, you learned how testosterone shapes the electric organ discharge (EOD) in *Brienomyrus brachyistius*, and you performed the implantation procedure yourself. Now it's time to analyze the data.

EOD recordings were captured daily over 5 days for animals implanted with either 11-ketotestosterone (11-kt) or a blank control. Each recording session is stored as a JSON file containing multiple trials. Our goal is to:

1. Load a raw EOD waveform and visualize it
2. Write a function that measures EOD duration from any recording file
3. Apply that function across all recording sessions to build a dataset
4. Visualize how EOD duration changes over time for each treatment group

## Loading EOD Recording Files in R

JSON files can be read in R using the `jsonlite` package. The `read_json()` function returns a list that maps directly onto the file structure.

```{r load-json, message=FALSE, warning=FALSE}
library(jsonlite)

eod_data <- read_json("data/eod_duration/BB_T48_DAY0_MAY7.json")

cat("Number of recordings:", length(eod_data), "\n")
```

Each element of `eod_data` is one recording trial, stored as a named list. Use `names(eod_data[[1]])` to see the available fields and `$` to access them.

::: {.callout-important title="Challenge 1: Explore the recording metadata (5 min)"}

Using the `eod_data` object already loaded, print the specimen number, species, temperature, sampling rate, and number of waveform samples for the first recording trial.

*Hint:* `names(eod_data[[1]])` shows the available fields.

:::: {.callout-tip title="Solution" collapse="true"}

```{r challenge-1-solution, eval=FALSE}
names(eod_data[[1]])
cat("Specimen:", eod_data[[1]]$specimenno, "\n")
cat("Species: ", eod_data[[1]]$species, "\n")
cat("Temp:    ", eod_data[[1]]$temp, "\n")
cat("Rate:    ", eod_data[[1]]$Rate, "\n")
cat("Samples: ", length(eod_data[[1]]$wave), "\n")
```

::::

:::

## Visualizing an EOD Waveform

We convert sample indices to milliseconds using the sampling rate stored in the file, then plot with `ggplot2`.

```{r plot-waveform, message=FALSE, warning=FALSE, fig.alt="Line plot of a single EOD waveform showing voltage in volts on the y-axis and time in milliseconds on the x-axis. The waveform has a characteristic biphasic shape with a positive P1 peak followed by a negative P2 trough.", fig.cap="A single raw EOD waveform recorded from *B. brachyistius*. The biphasic waveform is characteristic of mormyrid electric fish."}
library(ggplot2)

rate_hz <- eod_data[[1]]$Rate
wave <- unlist(eod_data[[1]]$wave)
p1_idx <- which.max(wave)
time_ms <- (seq_along(wave) - p1_idx) / rate_hz * 1000

waveform_df <- data.frame(time_ms = time_ms, voltage = wave)

ggplot(waveform_df, aes(x = time_ms, y = voltage)) +
  geom_line(linewidth = 0.6) +
  coord_cartesian(xlim = c(-2, 2)) +
  theme_classic() +
  labs(
    x = "Time (ms)",
    y = "Voltage (V)",
    title = paste("Raw EOD waveform —", eod_data[[1]]$specimenno, "Day 0")
  )
```


::: {.callout-important title="Challenge 2: Compare waveforms across recording sessions (5 min)"}

The Day 0 waveform for BB48 is already loaded as `eod_data`. Now load the Day 1 recording (`data/eod_duration/BB_T48_DAY1_MAY8.json`) and plot both waveforms together. Do the shapes look similar?

:::: {.callout-tip title="Solution" collapse="true"}

```{r challenge-2-solution, eval=FALSE}
eod_day1 <- read_json("data/eod_duration/BB_T48_DAY1_MAY8.json")
wave1     <- unlist(eod_day1[[1]]$wave)
p1_idx1   <- which.max(wave1)
time1_ms  <- (seq_along(wave1) - p1_idx1) / eod_day1[[1]]$Rate * 1000

# Normalize each waveform to peak-to-peak = 1 so amplitudes are comparable
wave0_norm <- wave  / (max(wave)  - min(wave))
wave1_norm <- wave1 / (max(wave1) - min(wave1))

df_compare <- rbind(
  data.frame(time_ms = time_ms,  voltage = wave0_norm, session = "Day 0"),
  data.frame(time_ms = time1_ms, voltage = wave1_norm, session = "Day 1")
)

ggplot(df_compare, aes(x = time_ms, y = voltage, color = session)) +
  geom_line(linewidth = 0.6) +
  coord_cartesian(xlim = c(-2, 2)) +
  theme_classic() +
  scale_colour_manual(values = c("Day 0" = "grey50", "Day 1" = "tomato")) +
  labs(
    x = "Time (ms)", y = "Normalized voltage",
    title = "BB48 EOD waveforms — Day 0 vs Day 1",
    color = NULL
  )
```

These are the first two days of recording for BB48, before treatment has had time to act. The waveform shapes should be very similar.

::::

:::

## Measuring EOD Duration

EOD duration is not read directly from the file — it must be calculated from the waveform. The standard approach is:

1. **Normalize** the waveform so that the peak-to-peak amplitude = 1
2. **Apply a threshold** (typically 10% of the maximum absolute value) to define where the EOD "starts" and "ends"
3. **Calculate duration** as the time elapsed between the first and last threshold crossings

```{r measure-duration, warning=FALSE}
# 1. Normalize to peak-to-peak = 1
wave_norm <- wave / (max(wave) - min(wave))

# 2. Apply 10% threshold
threshold <- 0.1 * max(abs(wave_norm))
above_thresh <- which(abs(wave_norm) > threshold)

start_idx <- min(above_thresh)
end_idx <- max(above_thresh)

# 3. Duration in milliseconds
duration_ms <- (end_idx - start_idx) / rate_hz * 1000
cat("EOD duration (10% threshold):", round(duration_ms, 3), "ms\n")
```

We can visualize the threshold on the normalized waveform to see exactly which part of the signal is being measured:

```{r plot-threshold, message=FALSE, warning=FALSE, fig.alt="Normalized EOD waveform with horizontal dashed lines at +10% and -10% of the peak amplitude. Vertical dashed lines mark the measured start and end of the EOD, highlighting the duration window.", fig.cap="Threshold-based EOD duration measurement on the normalized waveform."}
norm_df <- data.frame(time_ms = time_ms, voltage = wave_norm)

ggplot(norm_df, aes(x = time_ms, y = voltage)) +
  geom_line(linewidth = 0.6) +
  geom_hline(
    yintercept = c(threshold, -threshold),
    linetype = "dashed", color = "steelblue", linewidth = 0.5
  ) +
  geom_vline(
    xintercept = c(time_ms[start_idx], time_ms[end_idx]),
    linetype = "dashed", color = "tomato", linewidth = 0.5
  ) +
  coord_cartesian(xlim = c(-2, 2)) +
  theme_classic() +
  labs(
    x     = "Time (ms)",
    y     = "Normalized voltage",
    title = "EOD duration measurement (10% threshold)"
  )
```

::: {.callout-important title="Challenge 3: Sensitivity of the threshold (5 min)"}

Repeat the duration calculation using a **20% threshold** instead of 10%. How does the measured duration change, and why?

This is important: the choice of threshold is an **analytical decision** that must be reported in any publication.

:::: {.callout-tip title="Solution" collapse="true"}

```{r challenge-3-solution, eval=FALSE}
threshold_20 <- 0.2 * max(abs(wave_norm))
above_20 <- which(abs(wave_norm) > threshold_20)

duration_20ms <- (max(above_20) - min(above_20)) / rate_hz * 1000
cat("Duration at 20% threshold:", round(duration_20ms, 3), "ms\n")
cat("Duration at 10% threshold:", round(duration_ms, 3), "ms\n")
```

A higher threshold excludes the low-amplitude tails of the waveform, making the measured duration shorter. Neither threshold is "wrong" — but the choice must be consistent across all animals in a study and clearly reported.

::::

:::

## Building a Dataset Across All Recording Sessions

Now that we know how to measure EOD duration from a single file, we can apply the same logic to every recording session in the study. Rather than computing durations by hand for each file, we write a function and let R do the work.

The study metadata is stored in `recordings.csv`, which maps each JSON file to its animal, day, and treatment group:

```{r load-metadata, message=FALSE, warning=FALSE, eval=file.exists("data/eod_duration/all_surviving_recordings.csv")}
library(dplyr)

metadata <- read.csv("data/eod_duration/all_surviving_recordings.csv")
metadata$Treatment <- as.factor(metadata$Treatment)

metadata
```

We write a small function that takes a file path, loads the JSON, computes EOD duration for every trial, and returns the mean:

```{r duration-function, message=FALSE, warning=FALSE, eval=file.exists("data/eod_duration/all_surviving_recordings.csv")}
mean_eod_duration <- function(file_path, threshold = 0.1) {
  recs <- read_json(file_path)
  durations <- sapply(recs, function(rec) {
    w <- unlist(rec$wave)
    rate <- rec$Rate
    w_norm <- w / (max(w) - min(w))
    thresh <- threshold * max(abs(w_norm))
    above <- which(abs(w_norm) > thresh)
    (max(above) - min(above)) / rate * 1000
  })
  mean(durations)
}
```

We then apply this function to every row in the metadata table:

```{r compute-durations, message=FALSE, warning=FALSE, eval=file.exists("data/eod_duration/all_surviving_recordings.csv")}
metadata$EOD <- sapply(
  file.path("data/eod_duration", metadata$file),
  mean_eod_duration
)

head(metadata)
```

### Longitudinal trajectory

With the full dataset assembled, we can visualize how EOD duration changes across the 5-day experiment for each treatment group. Points show the group mean; error bars show ±1 standard deviation.

```{r longitudinal-plot, message=FALSE, warning=FALSE, eval=file.exists("data/eod_duration/all_surviving_recordings.csv"), fig.alt="Line plot showing mean EOD duration in milliseconds on the y-axis and experimental day on the x-axis, with separate colored lines for the 11-kt and control groups. Error bars represent one standard deviation.", fig.cap="Mean EOD duration (±SD) across experimental days by treatment group."}
ggplot(metadata, aes(x = day, y = EOD, group = Treatment, color = Treatment)) +
  stat_summary(
    fun = "mean",
    geom = "pointrange",
    fun.max = function(x) mean(x) + sd(x),
    fun.min = function(x) mean(x) - sd(x),
    position = position_dodge(width = 0.2, preserve = "total"),
    size = 0.7,
    fatten = 0.5
  ) +
  scale_x_continuous(name = "Day", breaks = 0:4) +
  scale_y_continuous(
    name   = "EOD duration (ms)",
    breaks = scales::breaks_pretty(10)
  ) +
  scale_colour_manual(values = c("11-kt" = "tomato", "control" = "grey40")) +
  theme_classic() +
  theme(
    axis.text    = element_text(size = 9),
    legend.text  = element_text(size = 8),
    legend.title = element_text(size = 9)
  )
```


## Visualizing Waveform Shape Change by Individual

Looking at the raw waveforms makes the treatment effect intuitive. Here we load the Day 0 and Day 4 recordings for each fish, normalize every waveform to a peak-to-peak amplitude of 1, and centre on the P1 peak. Overlaying the two traces for each animal shows directly how the waveform lengthens — or stays the same — depending on treatment.

```{r waveform-by-subject, message=FALSE, warning=FALSE, eval=file.exists("data/eod_duration/all_surviving_recordings.csv"), fig.alt="Four faceted panels, one per fish. Each panel shows two overlaid normalized EOD waveforms: Day 0 in grey and Day 4 in red, centred on the P1 peak. The two 11-kt fish show a clear rightward extension of the waveform by Day 4; the two control fish waveforms are nearly identical across days.", fig.cap="Day 0 (grey) and Day 4 (red) EOD waveforms for each individual fish, normalized to peak-to-peak amplitude and centred on P1. Waveforms from the 11-kt-treated fish (BB48, BB49INJ) show a lengthened P2 phase by Day 4; control fish (BB50, BB51INJ) are largely unchanged."}
load_waveform_norm <- function(file_path) {
  recs <- read_json(file_path)
  rec <- recs[[1]]
  w <- unlist(rec$wave)
  rate <- rec$Rate
  p1_idx <- which.max(w)
  w_norm <- w / (max(w) - min(w))
  data.frame(
    time_ms      = (seq_along(w) - p1_idx) / rate * 1000,
    voltage_norm = w_norm
  )
}

d0d2 <- metadata %>% filter(day %in% c(0, 4))

waveform_overlay <- bind_rows(lapply(seq_len(nrow(d0d2)), function(i) {
  fp <- file.path("data/eod_duration", d0d2$file[i])
  df <- load_waveform_norm(fp)
  df$fish_label <- paste0(d0d2$specimenno[i], " (", d0d2$Treatment[i], ")")
  df$Day <- paste0("Day ", d0d2$day[i])
  df
}))

ggplot(waveform_overlay, aes(x = time_ms, y = voltage_norm, color = Day)) +
  geom_line(linewidth = 0.6, alpha = 0.85) +
  coord_cartesian(xlim = c(-1.5, 3)) +
  facet_wrap(~fish_label, ncol = 2) +
  scale_colour_manual(values = c("Day 0" = "grey50", "Day 4" = "tomato")) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "top"
  ) +
  labs(
    x     = "Time (ms)",
    y     = "Normalized voltage",
    color = NULL,
    title = "EOD waveforms by individual: Day 0 vs Day 4"
  )
```
::: {.callout-tip title="Key Points"}

- `jsonlite::read_json()` loads JSON files as nested R lists; each recording trial is a named list with fields like `wave`, `Rate`, `temp`, and `specimenno`
- EOD duration is defined by a threshold criterion applied to the normalized waveform — the choice of threshold is an analytical decision that must be reported
- A metadata CSV that maps files to experimental conditions lets you build a tidy dataset by applying a duration function across all recording sessions

:::
