---
title: 'Normalization, Dimensionality Reduction, and Clustering'
teaching: 45
exercises: 10
---

::: {.callout-note title="Questions"}

- Why do we need to normalize single-nucleus RNA-seq count data?
- What is principal component analysis, and why is it used before clustering?
- How are cell clusters identified computationally, and what does "resolution" control?
- How do we distinguish biologically meaningful clusters from technical artifacts?

:::

::: {.callout-tip title="Objectives"}

- Merge individual per-sample Seurat objects and annotate cells with experimental condition metadata.
- Normalize count data using SCTransform to remove sequencing-depth variation.
- Interpret a PCA elbow plot to select the number of principal components for downstream analysis.
- Identify cell clusters using a graph-based Louvain algorithm.
- Visualize cells in UMAP space and color by cluster, sample, and experimental condition.

:::

## Setup

```{r load-libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(glmGamPoi)
library(presto)
library(clusterProfiler)
library(GO.db)
```

## Load the data

We reload the raw 10X count matrices for each midbrain sample, following the same approach used in the QC episode.

```{r load-data, message=FALSE, warning=FALSE}
tissue <- "EO"
sample_names <- c("BB48", "BB49INJ", "BB50", "BB50INJ")

# Empty list to store one Seurat object per sample
list_seurat <- list()

for (sample in sample_names) {
  data_dir <- file.path("data/ssrnaseq_data", tissue, sample)

  seurat_data <- Read10X(data.dir = data_dir)
  seurat_obj <- CreateSeuratObject(
    counts = seurat_data,
    min.features = 1,
    project = paste0(tissue, "_", sample)
  )

  list_seurat[[sample]] <- seurat_obj
}
```

## Merge samples and annotate experimental conditions

Before normalizing, we merge all four samples into a single Seurat object and immediately record which experimental condition each cell belongs to. This metadata travels with every cell through all downstream steps and is essential for interpreting your results.

```{r merge-and-annotate, message=FALSE, warning=FALSE}
# Merge into one object; add.cell.id prefixes each barcode with its sample name
# to prevent collisions across samples
merged_seurat <- merge(
  x = list_seurat[[1]],
  y = list_seurat[2:length(list_seurat)],
  add.cell.id = names(list_seurat)
)

# In Seurat v5, layers must be joined before downstream analysis
merged_seurat <- JoinLayers(merged_seurat)

# Assign experimental condition to each cell based on orig.ident
# BB48 and BB49INJ were implanted with 11-ketotestosterone (11kt) for 4 days
# BB50 and BB50INJ received a vehicle (cocoa butter) implant for 4 days
condition_map <- c(
  "EO_BB48"    = "11kt",
  "EO_BB49INJ" = "11kt",
  "EO_BB50"    = "vehicle",
  "EO_BB50INJ" = "vehicle"
)

merged_seurat$condition <- unname(condition_map[merged_seurat$orig.ident])

# Inspect the metadata to confirm the assignment looks right
head(merged_seurat@meta.data)
```

::: {.callout-note}

## Experimental design: 11-KT vs. vehicle

This dataset captures two biological conditions:

- **11kt**: Animals received a slow-release 11-ketotestosterone implant. 11-KT is the dominant androgen in teleost fish and is known to lengthen the electric organ discharge (EOD) duration.
- **vehicle**: Animals received an inert cocoa butter implant with no steroid.

Storing `condition` as metadata now means we can color UMAP plots, run differential expression tests, and check for batch effects throughout the rest of the analysis with a single column name.

:::

## Normalization with SCTransform

Raw UMI counts vary across cells partly because of real biology (some cell types transcribe more RNA) and partly because of technical noise (cells were captured and sequenced at different depths). Normalization removes the technical component so that downstream comparisons reflect biology.

We use **SCTransform**, which fits a regularized negative binomial regression model to each gene. The residuals from this model are variance-stabilized and depth-corrected, and they replace the raw counts as the working expression matrix (stored in a new `SCT` assay).

```{r sctransform, message=FALSE, warning=FALSE}
merged_seurat <- SCTransform(merged_seurat, verbose = FALSE)
```

::: {.callout-note}

## What about mitochondrial gene regression?

Many single-cell tutorials (including the CCR tutorial this lesson follows) regress out the percentage of mitochondrial reads (`percent.mt`) during SCTransform. High mitochondrial content often flags dying or damaged cells.

In mormyrid fish, mitochondrial genes do **not** follow the standard `^mt-` prefix used in mouse and human datasets. Before regressing, you would need to identify the correct prefix in your feature table (`features.tsv.gz`) and recalculate `PercentageFeatureSet()` accordingly. We skip this step here to keep the lesson focused — it is worth revisiting for a production analysis.

:::

::: {.callout-important}

## Inspect the SCT assay

After running SCTransform, the Seurat object now has two assays: `RNA` (raw counts) and `SCT` (normalized residuals).

1. How many variable features did SCTransform select? (*Hint:* `length(VariableFeatures(merged_seurat))`)
2. How does the `SCT` data slot differ from the `RNA` counts slot? (*Hint:* compare `merged_seurat[["RNA"]]$counts[1:5, 1:3]` with `merged_seurat[["SCT"]]$scale.data[1:5, 1:3]`)

:::: {.callout-tip title="Solution" collapse="true"}

```r
# Number of variable features
length(VariableFeatures(merged_seurat))
# SCTransform typically selects ~3,000 variable features by default

# Raw RNA counts (integers, depth-variable)
merged_seurat[["RNA"]]$counts[1:5, 1:3]

# SCT scaled data (continuous residuals, variance-stabilized)
merged_seurat[["SCT"]]$scale.data[1:5, 1:3]
```

The `SCT` scale data values are no longer integers — they are Pearson residuals centered near zero. Genes expressed at higher levels than expected for a cell's sequencing depth get positive residuals; lower-than-expected expression gives negative residuals.

::::

:::

## Principal component analysis (PCA)

Even after selecting ~3,000 variable features, analyzing all genes simultaneously is computationally expensive and noisy. PCA compresses this high-dimensional space into a small number of **principal components** (PCs), each a weighted combination of genes that captures a distinct axis of variation in the data.

```{r run-pca, message=FALSE, warning=FALSE}
merged_seurat <- RunPCA(merged_seurat, verbose = FALSE, assay = "SCT")
```

### Visualize PC loadings

`DimHeatmap` shows which genes drive each PC. Real biological signal tends to produce clean, polarized heatmaps; noisy PCs show less structure.

```{r dim-heatmap, fig.width=10, fig.height=12, message=FALSE, warning=FALSE}
DimHeatmap(merged_seurat, dims = 1:9, cells = 500, balanced = TRUE, ncol = 3)
```

### Elbow plot: how many PCs to use?

The elbow plot ranks PCs by the amount of variance they explain. The goal is to include PCs that carry biological signal while excluding those that are dominated by noise. Look for the "elbow" — the point where the curve flattens — and choose a cutoff just before it.

```{r elbow-plot, message=FALSE, warning=FALSE}
ElbowPlot(merged_seurat, ndims = 40)
```

::: {.callout-note}

## Reading the elbow plot

There is no single correct answer. Common practice:

- Use enough PCs that the elbow is comfortably included (typically 15–30 for most datasets).
- Err on the side of more PCs — including a few extra noisy PCs rarely hurts, but dropping informative ones can merge biologically distinct clusters.
- For this dataset we use **25 PCs** in the steps below.

:::

::: {.callout-important}

## Choose your own PC cutoff

Look at the elbow plot above.

1. Where do you think the "elbow" falls for this dataset?
2. Re-run `FindNeighbors` and `FindClusters` (next section) with `dims = 1:15` instead of `dims = 1:25`. Does the number of clusters change?

:::: {.callout-tip title="Solution" collapse="true"}

The elbow for this dataset typically falls around PC 20–30, where the curve visibly flattens. Using `dims = 1:15` may merge some clusters that are separated with more PCs, particularly for rarer cell types. There is no universally correct answer — the choice should be informed by biological expectations and reproducibility across a range of values.

::::

:::

## Graph-based clustering

Seurat clusters cells using a two-step approach:

1. **`FindNeighbors`**: Builds a *k*-nearest neighbor (kNN) graph in PCA space. Cells that are close together in PC space (i.e., have similar transcriptomes) are connected by edges.
2. **`FindClusters`**: Applies the Louvain algorithm to partition the graph into communities (clusters) by maximizing the number of within-cluster edges relative to what you would expect by chance.

```{r find-clusters, message=FALSE, warning=FALSE}
merged_seurat <- FindNeighbors(merged_seurat, dims = 1:25)
merged_seurat <- FindClusters(merged_seurat, resolution = 1)
```

::: {.callout-note}

## The `resolution` parameter

`resolution` controls how finely the graph is partitioned:

| Resolution | Effect |
|-----------|--------|
| Low (e.g., 0.1) | Fewer, larger clusters — broad cell types |
| Medium (e.g., 0.5) | Intermediate granularity |
| High (e.g., 1.0–2.0) | Many small clusters — sub-types and states |

Start with a moderate resolution and adjust based on biology. Comparing your clusters to known cell-type markers is the only way to judge whether a given resolution is biologically meaningful.

:::

## UMAP visualization

UMAP (Uniform Manifold Approximation and Projection) maps the high-dimensional PCA representation into two dimensions for visualization. Cells that are transcriptomically similar end up close together on the UMAP. Unlike PCA, UMAP is non-linear, so distances between distant clusters are not directly comparable.

```{r run-umap, message=FALSE, warning=FALSE}
merged_seurat <- RunUMAP(merged_seurat, dims = 1:25)
```

```{r umap-plots, fig.width=14, fig.height=5, message=FALSE, warning=FALSE}
p1 <- DimPlot(merged_seurat,
  reduction = "umap",
  group.by = "seurat_clusters", label = TRUE, repel = TRUE
) +
  ggtitle("Cluster") +
  theme(legend.position = "none")

p2 <- DimPlot(merged_seurat,
  reduction = "umap",
  group.by = "orig.ident", alpha = 0.4
) +
  ggtitle("Sample")

p3 <- DimPlot(merged_seurat,
  reduction = "umap",
  group.by = "condition", alpha = 0.4,
  cols = c("11kt" = "#D55E00", "vehicle" = "#0072B2")
) +
  ggtitle("Condition (11kt vs. vehicle)")

p1 | p2 | p3
```

::: {.callout-note}

## What to look for in UMAP plots

**By cluster:** Are clusters well-separated, or do they blur into each other? Highly diffuse boundaries may indicate over-clustering.

**By sample:** Do all four samples contribute cells to each cluster? If one cluster is dominated by a single sample, it may reflect a technical artifact (e.g., ambient RNA contamination or a doublet population) rather than a real cell type.

**By condition:** Do 11kt-treated and vehicle cells intermix within clusters, or are some clusters exclusive to one condition? Intermixing suggests shared cell types are present in both groups; condition-specific clusters indicate populations that appear or disappear with hormone treatment.

:::

::: {.callout-important}

## Explore clustering resolution

Re-run `FindClusters` at two different resolutions — `resolution = 0.1` and `resolution = 1.0` — and regenerate the UMAP colored by cluster for each.

1. How many clusters does each resolution produce?
2. Do the broader clusters at low resolution correspond to merged versions of the fine-grained clusters at high resolution, or are they arranged differently?
3. Which resolution do you think is more biologically interpretable for this dataset?

:::: {.callout-tip title="Solution" collapse="true"}

```r
# Low resolution
merged_seurat <- FindClusters(merged_seurat, resolution = 0.1)
DimPlot(merged_seurat, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE) + ggtitle("Resolution = 0.1")

# High resolution
merged_seurat <- FindClusters(merged_seurat, resolution = 1.0)
DimPlot(merged_seurat, reduction = "umap", group.by = "seurat_clusters",
        label = TRUE) + ggtitle("Resolution = 1.0")
```

Low resolution (0.1) typically yields ~5–8 clusters corresponding to broad cell classes (neurons, glia, vascular cells, etc.). High resolution (1.0) can produce 20+ clusters, splitting these classes into subtypes and activation states. The "correct" resolution depends on your biological question — broad cell typing needs lower resolution; identifying a rare hormone-responsive subpopulation may require higher resolution.

::::

:::

## Finding cluster biomarkers

Now that cells are clustered, we can ask: **which genes best define each cluster?** These marker genes are the bridge between an unlabeled cluster number and a biological cell-type identity.

We use `FindAllMarkers()`, which runs a statistical test (Wilcoxon rank-sum by default) comparing each cluster against all other cells. Because we used SCTransform, we first call `PrepSCTFindMarkers()` to re-correct the counts to a common sequencing depth across samples — this is important when your data spans multiple samples (as ours does).
```{r prep-markers, message=FALSE, warning=FALSE}
# Required before FindAllMarkers when SCTransform was used with multiple samples
merged_seurat <- PrepSCTFindMarkers(merged_seurat, verbose = FALSE)
```

```{r find-markers, message=FALSE, warning=FALSE}
# Find markers for every cluster vs. all remaining cells
# only.pos = TRUE keeps only upregulated markers (genes enriched in the cluster)
all_markers <- FindAllMarkers(
  merged_seurat,
  only.pos = TRUE,
  verbose = FALSE
)

# Sort by cluster, then by effect size (avg_log2FC) and significance
all_markers <- all_markers %>%
  arrange(cluster, desc(avg_log2FC), desc(-p_val_adj))

head(all_markers, 20)
```

::: {.callout-note}

## Interpreting the marker table

| Column | Meaning |
|--------|---------|
| `cluster` | Which cluster this marker defines |
| `gene` | Gene name |
| `avg_log2FC` | Log₂ fold-change vs. all other cells — larger = more specific |
| `p_val_adj` | Benjamini–Hochberg adjusted p-value |
| `pct.1` | Fraction of cells in this cluster expressing the gene |
| `pct.2` | Fraction of all other cells expressing the gene |

A good marker has high `avg_log2FC`, low `p_val_adj`, high `pct.1`, and low `pct.2`. The ratio `pct.1 / pct.2` is a useful intuitive measure of specificity.

:::

### Top markers per cluster

```{r top-markers}
# Pull the top 5 markers per cluster by log2FC
top5 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 5) %>%
  ungroup()

top5
```

### Heatmap of top markers

A heatmap lets you visually assess whether the top markers are truly specific to their cluster or broadly expressed.

```{r marker-heatmap, fig.width=12, fig.height=8, message=FALSE, warning=FALSE}
# Use top 10 per cluster for the heatmap
top10 <- all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 10) %>%
  ungroup()

DoHeatmap(merged_seurat, features = top10$gene) + NoLegend()
```

### Feature plots for selected markers

`FeaturePlot` overlays gene expression on the UMAP, making it easy to see whether a marker is truly restricted to one cluster or bleeds into neighboring ones.

```{r feature-plots, fig.width=10, fig.height=4, message=FALSE, warning=FALSE}
# Grab the single top marker for the first three clusters as an example
example_genes <- all_markers %>%
  group_by(cluster) %>%
  slice_max(avg_log2FC, n = 1) %>%
  ungroup() %>%
  slice_head(n = 3) %>%
  pull(gene)

FeaturePlot(merged_seurat, features = "LOC125742960", ncol = 3)
```

::: {.callout-important}

## Identify a cluster of interest

Look at the `top5` marker table and the UMAP.

1. Pick one cluster whose top markers look biologically interesting to you. Search for those gene names in a vertebrate gene database (e.g., NCBI Gene, Ensembl, or zfin.org for teleost fish) — what cell type do they suggest?
2. Use `VlnPlot()` to show expression of your chosen marker split by `condition`. Is the marker expressed differently in 11kt-treated vs. vehicle cells within that cluster?

```r
VlnPlot(merged_seurat, features = "YOUR_GENE_HERE",
        group.by = "seurat_clusters", split.by = "condition")
```

:::: {.callout-tip title="Solution" collapse="true"}

There is no single correct answer — this exercise is open-ended by design. Common cell types found in teleost midbrain include:

- **Neurons**: express *snap25a/b*, *syt1a/b*, *elavl3/4*
- **Oligodendrocytes / myelinating glia**: express *mbpa*, *plp1a*, *cnp*
- **Astrocytes**: express *gfap*, *aldh1l1*, *slc1a2*
- **Microglia**: express *ptprc* (CD45), *csf1ra*, *aif1l*
- **Endothelial cells**: express *pecam1*, *cdh5*, *kdr*

If a marker shows differential expression between conditions within a cluster, that cluster may contain cells responding to 11-KT treatment — a hypothesis worth testing with formal differential expression analysis in a later episode.

::::

:::

## Annotating markers with gene descriptions

Looking up `LOC125746627` in a gene database every time you encounter an unknown gene quickly becomes a bottleneck. A faster approach is to join the full genome annotation — the GTF file downloaded from NCBI — directly onto your marker table so every row already carries a human-readable description.

### Parse the GTF annotation

The GTF file for *Brienomyrus brachyistius* (NCBI RefSeq accession GCF_023856365.1) stores gene product names in the `product` attribute of each transcript feature. We read the file, extract `gene_id` and `product` with regular expressions, and keep one description per gene.

```{r parse-gtf, message=FALSE, warning=FALSE}
# Read the GTF, skipping comment lines that begin with "#"
gtf_raw <- read_tsv(
  "data/ssrnaseq_data/genes.gtf.gz",
  comment        = "#",
  col_names      = FALSE,
  col_types      = "ccciicccc",
  show_col_types = FALSE
)

# Keep only rows that carry a "product" attribute (mRNA / transcript rows)
# then extract gene_id and product via regex
gene_desc <- gtf_raw |>
  filter(str_detect(X9, 'product "')) |>
  mutate(
    gene_id = str_extract(X9, 'gene_id "([^"]+)"', group = 1),
    product = str_extract(X9, 'product "([^"]+)"',  group = 1)
  ) |>
  dplyr::select(gene_id, product) |>
  distinct(gene_id, .keep_all = TRUE)  # one description per gene

nrow(gene_desc)
head(gene_desc)
```

### Join descriptions onto the marker table

```{r annotate-markers, message=FALSE, warning=FALSE}
annotated_markers <- all_markers |>
  left_join(gene_desc, by = c("gene" = "gene_id"))

# Top 5 per cluster with descriptions
annotated_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5) |>
  ungroup() |>
  dplyr::select(cluster, gene, product, avg_log2FC, p_val_adj, pct.1, pct.2)
```

::: {.callout-note}

## Why are so many genes called "LOC######"?

NCBI assigns `LOC` names (e.g., `LOC125746627`) as **provisional gene symbols** when a gene has been computationally predicted but lacks an official name from a nomenclature committee. The number is the NCBI Gene ID. These genes *do* have functional annotations — you just have to look in the `product` field rather than the gene name.

Three things you'll commonly see in the product description:

| Product text | What it means |
|---|---|
| `"synaptosome associated protein 25b"` | Well-characterized ortholog of a known vertebrate gene |
| `"potassium voltage-gated channel…-like"` | Computationally inferred function from sequence similarity |
| `"uncharacterized protein"` | No confident functional inference yet |

The `-like` and `uncharacterized` entries are not dead ends — they tell you the gene family (e.g., "voltage-gated channel") even if the exact paralog is uncertain. For *B. brachyistius*, a non-model organism with a newly sequenced genome, this level of annotation is typical and improves over time as more papers are published.

:::

::: {.callout-important}

## Propose cell-type labels from the annotated table

Scan the annotated top-5 marker table above. For each cluster:

1. List the product descriptions of its top markers.
2. Based on those descriptions, propose a cell-type label (e.g., "neuron", "oligodendrocyte").
3. Are there any clusters where all five top markers are "uncharacterized"? What would you do next to annotate those?

Use the following teleost brain cell-type reference as a starting point:

| Cell type | Canonical marker genes |
|-----------|----------------------|
| Neuron | *snap25a/b*, *syt1a/b*, *elavl3/4* |
| Oligodendrocyte | *mbpa*, *plp1a*, *cnp* |
| Astrocyte | *gfap*, *aldh1l1*, *slc1a2* |
| Microglia | *ptprc*, *csf1ra*, *aif1l* |
| Endothelial | *pecam1*, *cdh5*, *kdr* |

:::: {.callout-tip title="Solution" collapse="true"}

There is no single correct answer — it depends on the clustering resolution and the specific dataset. General strategy:

- Clusters whose top markers include descriptions like "synaptosome-associated", "vesicle", "voltage-gated channel", or "neurexin" → **neurons**
- Clusters with "myelin", "proteolipid protein", "oligodendrocyte" → **oligodendrocytes / myelinating glia**
- Clusters with "glial fibrillary", "glutamate transporter", "aquaporin" → **astrocytes**
- Clusters with "complement receptor", "colony stimulating factor receptor" → **microglia**

For clusters where all top markers are "uncharacterized", useful next steps include:
- Examining lower-ranked markers that may have named orthologs
- BLASTing the LOC protein sequence against zebrafish or human proteomes
- Checking whether the cluster is condition-specific (a possible artifact or rare reactive state)

::::

:::

## Automated cluster annotation with Gene Ontology

Manually looking up every LOC gene is slow. A faster and more systematic approach is **Gene Ontology (GO) enrichment analysis**: for each cluster, we ask which GO biological-process terms are statistically over-represented among the cluster's marker genes compared to all tested genes. Clusters enriched for terms like *chemical synaptic transmission* likely contain neurons; clusters enriched for *myelination* are probably oligodendrocytes.

NCBI provides a GO annotation file (GAF format) for *B. brachyistius* that maps each gene to GO terms. We use that file together with `clusterProfiler` to run enrichment for all clusters at once.

### Parse the GO annotation file

The GAF file has one gene–GO association per row. We need two tables for `clusterProfiler`:
- **TERM2GENE**: links each GO ID to the gene symbols annotated with it
- **TERM2NAME**: links each GO ID to its human-readable name (retrieved from the `GO.db` package)

We restrict to **Biological Process** terms (aspect `"P"`) because those best reflect cell identity.

```{r parse-gaf, message=FALSE, warning=FALSE}
gaf <- read_tsv(
  "data/ssrnaseq_data/gene_ontology.gaf.gz",
  comment       = "!",
  col_names     = FALSE,
  col_types     = cols(.default = "c"),
  show_col_types = FALSE
)

# Column 3 = gene symbol (matches Seurat gene names)
# Column 5 = GO ID
# Column 9 = Aspect (P = Biological Process, F = Molecular Function,
#                    C = Cellular Component)
term2gene_bp <- gaf |>
  filter(X9 == "P") |>
  dplyr::select(go_id = X5, gene = X3) |>
  distinct()

# Retrieve human-readable GO term names from GO.db
# Term() returns a named character vector: names = GO IDs, values = term names
go_ids       <- unique(term2gene_bp$go_id)
term2name_bp <- tibble(
  go_id     = go_ids,
  term_name = Term(go_ids)
)

cat("Gene–GO associations (biological process):", nrow(term2gene_bp), "\n")
```

### Run GO enrichment for every cluster

`compareCluster()` runs `enricher()` on each cluster's marker genes in a single call and returns a tidy result ready for visualization. We use only statistically significant markers (adjusted *p* < 0.05) as the gene list, and all genes tested by `FindAllMarkers` as the background universe.

```{r go-enrichment, message=FALSE, warning=FALSE}
# Per-cluster gene lists (significant markers only)
sig_markers  <- annotated_markers |> filter(p_val_adj < 0.05)
cluster_genes <- split(sig_markers$gene, sig_markers$cluster)

# Background: every gene tested across all clusters
universe_genes <- unique(all_markers$gene)

go_compare <- compareCluster(
  geneCluster  = cluster_genes,
  fun          = "enricher",
  universe     = universe_genes,
  TERM2GENE    = term2gene_bp,
  TERM2NAME    = term2name_bp,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize    = 5,
  maxGSSize    = 500
)

dotplot(go_compare, showCategory = 5, font.size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Top enriched GO biological-process terms per cluster")
```

Each dot represents one enriched GO term. Dot size encodes the number of cluster markers that hit that term; colour encodes the adjusted *p*-value.

::: {.callout-note}

## Why use a background universe?

GO enrichment tests whether a term appears more often in your gene list than you would expect **by chance**, given how many genes in the genome are annotated to that term. Without a universe, every term looks enriched simply because the genome is large. Here we use all genes that `FindAllMarkers` tested — genes actually expressed in the dataset — rather than the full genome, which gives a more conservative and biologically meaningful result.

:::

### Assign automatic cluster labels

We can go one step further and convert GO enrichment results into cell-type labels. The idea is simple: each major brain cell type has a small set of highly discriminating GO terms. We count how many of those terms are enriched in each cluster and assign the best-matching label.

```{r auto-annotate, message=FALSE, warning=FALSE}
# Discriminating GO terms for major electric organ cell types
cell_type_signatures <- list(
  Electrocyte = c("GO:0006814",  # sodium ion transport
                  "GO:0006813",  # potassium ion transport
                  "GO:0019228",  # neuronal action potential
                  "GO:0003012",  # muscle system process
                  "GO:0006936"), # muscle contraction
  Muscle      = c("GO:0007519",  # skeletal muscle tissue development
                  "GO:0060537",  # muscle tissue development
                  "GO:0014706",  # striated muscle tissue development
                  "GO:0030239",  # myofibril assembly
                  "GO:0045214"), # sarcomere organization
  Schwann     = c("GO:0042552",  # myelination
                  "GO:0007272",  # ensheathment of neurons
                  "GO:0008366"), # axon ensheathment
  Neuron      = c("GO:0007268",  # chemical synaptic transmission
                  "GO:0007271",  # synaptic transmission, cholinergic
                  "GO:0007269"), # neurotransmitter secretion
  Fibroblast  = c("GO:0030198",  # extracellular matrix organization
                  "GO:0030199",  # collagen fibril organization
                  "GO:0042060"), # wound healing
  Endothelial = c("GO:0001568",  # blood vessel development
                  "GO:0001944",  # vasculature development
                  "GO:0035924")  # response to VEGF stimulus
)

# Flat table of all enrichment results
go_table <- as.data.frame(go_compare)

# Score each cluster against each cell-type signature
scores <- expand.grid(
  cluster   = unique(go_table$Cluster),
  cell_type = names(cell_type_signatures),
  stringsAsFactors = FALSE
) |>
  rowwise() |>
  mutate(
    score = sum(cell_type_signatures[[cell_type]] %in%
                  go_table$ID[go_table$Cluster == cluster])
  ) |>
  ungroup()

# Best-match label (score = 0 → "Unknown")
auto_labels <- scores |>
  group_by(cluster) |>
  slice_max(score, n = 1, with_ties = FALSE) |>
  mutate(label = if_else(score == 0, "Unknown", cell_type)) |>
  dplyr::select(cluster, label, score) |>
  arrange(as.integer(cluster))

auto_labels

# Store labels in the Seurat object
# unname() is required: subsetting a named vector preserves the index names
# (cluster IDs), not cell barcodes — Seurat needs an unnamed vector so it
# assigns values positionally.
label_vec <- setNames(auto_labels$label, auto_labels$cluster)
merged_seurat$cell_type_go <-
  unname(label_vec[as.character(merged_seurat$seurat_clusters)])

DimPlot(merged_seurat, group.by = "cell_type_go", label = TRUE, repel = TRUE) +
  ggtitle("GO-based automated cluster annotation") +
  NoLegend()
```

::: {.callout-important}

## Compare GO labels with your manual annotation

1. Look at the `auto_labels` table above. Do the GO-based labels match the cell types you proposed from the marker gene product descriptions in the previous challenge?
2. Are there any clusters labelled **"Unknown"**? For those clusters, what additional steps might help you assign a cell type?
   - Hint: re-run the GO enrichment restricting `filter(X9 == "F")` (Molecular Function) or `"C"` (Cellular Component) instead of Biological Process — do those aspects provide any additional clues?
3. Electrocytes are the dominant cell type in the electric organ and are derived from muscle. Does the UMAP reflect that — are electrocyte and muscle clusters the largest, and are they positioned close together?

:::: {.callout-tip title="Solution" collapse="true"}

There is no single correct answer. General expectations for a teleost electric organ dataset:

- One or more large clusters should receive the **Electrocyte** label, consistent with electrocytes being the numerically dominant cell type. Because electrocytes are myogenic (muscle-derived), **Muscle** clusters may appear nearby on the UMAP.
- At least one smaller cluster should be labelled **Schwann**, reflecting the dense myelinated innervation of the organ.
- **Neuron** clusters (electromotor nerve terminals) may be sparse or absent in a nuclear isolation prep, depending on how well small cholinergic terminals survive dissociation.
- **Fibroblast** and **Endothelial** clusters are typically small and compact, appearing on the UMAP periphery.
- **Unknown** clusters may represent rare cell types (e.g., pericytes, resident macrophages, progenitor cells) or electrocyte sub-states (e.g., innervated vs. non-innervated face gene expression) that lack specific GO annotations in a non-model species. For these, examining Molecular Function and Cellular Component terms, or BLASTing top marker sequences against zebrafish or human proteomes, are useful next steps.

The GO-based labels are a *starting point*, not ground truth. They should be treated as hypotheses and cross-validated against the canonical marker gene table from the previous challenge.

::::

:::

## Save the processed object

Save the normalized, clustered, and marker-annotated Seurat object for use in subsequent episodes.

```{r save-object, message=FALSE, warning=FALSE}
saveRDS(merged_seurat,
  file = file.path(
    "data/ssrnaseq_data", tissue,
    paste0(tissue, "_normalized_clustered.rds")
  )
)
```

::: {.callout-tip title="Key Points"}

- SCTransform normalizes single-nucleus count data by fitting a regularized negative binomial model per gene, producing Pearson residuals that correct for sequencing depth without discarding biological variation.
- PCA reduces the ~3,000 variable features to a compact set of principal components; the elbow plot guides the choice of how many PCs to retain.
- Graph-based clustering (`FindNeighbors` + `FindClusters`) groups cells by transcriptomic similarity in PCA space; the `resolution` parameter controls cluster granularity.
- UMAP provides a 2D visualization of cluster structure but distances between distant clusters should not be over-interpreted.
- Coloring the UMAP by experimental `condition` (11kt vs. vehicle) is an essential sanity check: shared cell types should be present in both conditions, while condition-specific clusters warrant further investigation.
- `PrepSCTFindMarkers()` + `FindAllMarkers()` identifies genes that best define each cluster; markers guide biological interpretation and cell-type annotation.
- The NCBI RefSeq GTF `product` field provides human-readable descriptions for LOC genes and is the primary tool for functional interpretation in non-model organism datasets.
- GO enrichment analysis (`clusterProfiler::compareCluster`) provides a systematic, automated starting point for cluster annotation by linking marker genes to biological-process terms; labels should be treated as hypotheses and cross-validated with canonical marker genes.

:::
