---
title: "Cell-Type Annotation and Differential Expression in the Electric Organ"
mode: bioinformatics
exercises: 90
optional: false
location: Loeb 156 (Breakroom)      # room shown in the schedule group heading, e.g. Loeb 160
---

::: {.callout-note title="Questions"}

- How do we turn unlabeled clusters into named cell types using curated marker genes?
- What anatomically distinguishes stalk electrocytes from face electrocytes, and how do we detect that split molecularly?
- Why is it statistically wrong to compare individual nuclei between treatment conditions, and what does pseudobulk fix?
- Does our snRNA-seq pipeline recover the gene-expression signatures of the androgen response previously found by bulk RNA-seq?

:::

::: {.callout-tip title="Objectives"}

- Load the post-Harmony checkpoint produced in Friday's preprocessing episode.
- Find cluster markers with `FindAllMarkers` and join NCBI gene-product descriptions for interpretation.
- Assign cell-type labels (electrocyte, muscle, Schwann, endothelial, motor neuron) using curated marker gene lists and `AddModuleScore`.
- Sub-classify electrocytes into stalk vs. face using PMCA paralog expression.
- Aggregate raw counts per (cell type × sample) and run DESeq2 with treatment as the contrast.
- Plot per-fish pseudobulk values for top DE genes and interpret the n=2-per-group result honestly.
- Cross-validate against the Losilla & Gallant 2025 bulk RNA-seq DEG table.

:::

## Introduction

Yesterday you took raw CellRanger output for four EO samples, normalized it, corrected for GEM-well batch with Harmony, and produced a clustered UMAP. The end state — saved as `data/checkpoints/02_harmony_clustered.rds` — is a Seurat object with unlabeled numbered clusters.

Today is the biological payoff. We assign cell-type identities to those clusters, sub-divide the electrocytes into stalk vs. face populations, and run the central statistical test of the experiment: **does 11-KT change gene expression in the electric organ, and if so, in which cells and which genes?**

The episode has four parts:

1. **Cluster markers.** What genes define each cluster, and what do they tell us about cell identity?
2. **Cell-type annotation.** Use curated marker gene lists and module scores to label every cluster.
3. **Stalk vs. face.** PMCA paralog expression separates innervated stalklets from the bulk of the electrocyte population.
4. **Differential expression.** Pseudobulk + DESeq2 contrast 11-KT vs. vehicle within each cell type; cross-validate against Losilla & Gallant 2025.

## Setup and load Friday's checkpoint

The checkpoint is the starting state regardless of whether you ran Friday's full pipeline. If you finished Friday, your in-memory object should already match; loading the checkpoint guarantees a known-good entry point.

```{r load-libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(glmGamPoi)
library(presto)
library(DESeq2)
library(limma)
library(ggrepel)
if (requireNamespace("speckle", quietly = TRUE)) library(speckle)
```

```{r load-checkpoint, message=FALSE, warning=FALSE}
merged_seurat <- readRDS("data/checkpoints/02_harmony_clustered.rds")
merged_seurat
```

## Find cluster markers

`FindAllMarkers` runs a Wilcoxon test for each cluster against all other cells, returning genes that distinguish each cluster from the rest. Before calling it on SCT-normalized data with multiple samples, we run `PrepSCTFindMarkers()` to re-correct counts to a common depth across samples.

```{r prep-markers, message=FALSE, warning=FALSE}
merged_seurat <- PrepSCTFindMarkers(merged_seurat, verbose = FALSE)
```

```{r find-markers, message=FALSE, warning=FALSE}
all_markers <- FindAllMarkers(
  merged_seurat,
  only.pos = TRUE,
  verbose  = FALSE
)

all_markers <- all_markers |>
  arrange(cluster, desc(avg_log2FC))

head(all_markers, 20)
```

### Annotate with gene-product descriptions

The Seurat row names mix curated gene symbols and `LOC#####` IDs. To make the marker table interpretable, we parse the *B. brachyistius* GTF and join the human-readable product description for each gene.

```{r parse-gtf, message=FALSE, warning=FALSE}
gtf_raw <- read_tsv(
  "data/ssrnaseq_data/genes.gtf.gz",
  comment        = "#",
  col_names      = FALSE,
  col_types      = "ccciicccc",
  show_col_types = FALSE
)

gene_desc <- gtf_raw |>
  filter(str_detect(X9, 'product "')) |>
  mutate(
    gene_id   = str_extract(X9, 'gene_id "([^"]+)"', group = 1),
    gene_name = str_extract(X9, 'gene "([^"]+)"', group = 1),
    product   = str_extract(X9, 'product "([^"]+)"', group = 1)
  ) |>
  dplyr::select(gene_id, gene_name, product) |>
  distinct(gene_id, .keep_all = TRUE)

# Build a combined lookup indexed by either gene_id (LOC ID) OR gene_name (symbol)
gene_lookup <- bind_rows(
  gene_desc |> transmute(key = gene_id, gene_name, product),
  gene_desc |>
    filter(!is.na(gene_name), gene_name != gene_id) |>
    transmute(key = gene_name, gene_name, product)
) |>
  distinct(key, .keep_all = TRUE)
```

```{r annotate-markers, message=FALSE, warning=FALSE}
annotated_markers <- all_markers |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key")
  )

annotated_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 3) |>
  ungroup() |>
  dplyr::select(cluster, gene, gene_name, product, avg_log2FC, p_val_adj) |>
  head(20)
```

::: {.callout-note title="Why so many `LOC#####`?"}

NCBI assigns `LOC` names as provisional symbols when a gene is computationally predicted but lacks an official name. The number is the NCBI Gene ID. The `product` field is where the function lives — "potassium voltage-gated channel...like" is a meaningful clue even when the gene name is `LOC125743019`.

For non-model genomes like *B. brachyistius*, this level of annotation is typical and improves over time as more papers are published.

:::

## Curated cell-type annotation

Generic GO enrichment performs poorly when one cell type dominates a tissue. Instead, we annotate using **curated marker gene lists** validated in published work on the mormyrid electric organ.

Two electrocyte markers are unambiguous across the literature:

- **scn4aa** (`LOC125742960`) — Nav1.4a, the principal electrogenic Na⁺ channel. Convergently recruited into electric organs across multiple independent electric-fish lineages (Gallant et al. 2014 *Science*; Thompson et al. 2018 *Curr Biol*).
- **kcna7a** (`LOC125743019`) — Kv1.7a, the delayed-rectifier K⁺ channel setting EOD repolarization kinetics (Losilla & Gallant 2025 *JEB*).

Electrocytes are myogenic in origin (derived from muscle), so muscle satellite cells (*pax7b*, myomesins) typically sit near electrocytes on the UMAP. Schwann cells, vascular endothelial cells, and a small motor neuron population round out the expected atlas.

```{r marker-gene-lists, message=FALSE, warning=FALSE}
marker_genes <- list(
  Electrocyte = c(
    # Sodium channels — Nav1.4 paralogs
    "LOC125742960", # scn4aa
    "scn4ab",
    # Potassium channels — repolarisation
    "LOC125743019", # kcna7a
    "kcna7", "kcnq1.2", "kcnq5a", "kcnn2",
    # Ion transport / electrocyte identity
    "cnga3a", "atp1b2b",
    # Cell-surface and structural identity
    "rrad", "igsf9ba", "robo2", "tenm1"
  ),
  Muscle = c(
    "pax7b", "pitx3", "myom2a", "myom2b",
    "hey2", "ezh2", "megf10", "mcama",
    "fgfr4", "tbx22"
  ),
  Schwann = c(
    "col28a1a", "smoc1", "prss12", "antxr1b",
    "numb", "olfml2ba", "col6a3", "sema3bl"
  ),
  Endothelial = c(
    "cdh5", "pecam1b", "kdr", "flt1",
    "clec14a", "pcdh12", "podxl", "adgrl4"
  ),
  Motorneuron = c(
    "mnx1",
    "LOC125712540", # isl-1
    "LOC125745880", # isl-1-like
    "LOC125704285", # isl-2a
    "LOC125706047", # isl-2a-like
    "chata", "slc18a3a",
    "elavl3", "rbfox3a",
    "nefma", "nefla", "neflb",
    "LOC125751823", "prph",
    "sv2a", "snap25a",
    "LOC125712969",
    "gap43", "nrn1b", "nrxn1a"
  )
)

# Filter to genes actually present in this dataset
marker_genes <- lapply(marker_genes, function(g) g[g %in% rownames(merged_seurat)])
print(sapply(marker_genes, length))

# Stalk sub-type marker: PMCA (plasma membrane Ca2+-ATPase) paralogs
# Baumann et al. 2025 — pan-PMCA protein enriched on stalklet membrane
pmca_genes <- c("atp2b1a", "atp2b2", "atp2b3b", "atp2b4")
pmca_genes <- pmca_genes[pmca_genes %in% rownames(merged_seurat)]
```

### Marker dot plot across clusters

A compact diagnostic — 3–4 most informative markers per cell type — projected as a dot plot across clusters. Dot size = fraction of cells expressing; color = mean expression.

```{r marker-dotplot, fig.width=15, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Dot plot showing curated cell-type marker gene expression across Seurat clusters, including PMCA stalk markers.", fig.cap="Canonical cell-type marker expression across EO clusters. scn4aa and scn4ab are electrocyte Na+ channels; pax7b marks muscle satellite cells; col28a1a marks Schwann cells; cdh5 marks endothelial cells; mnx1 and chata mark motor neurons. atp2b paralogs (PMCA) mark the stalk electrocyte sub-cluster."}
dot_genes <- c(
  "LOC125742960", "scn4ab", "atp1b2b", "rrad",
  pmca_genes,
  "pax7b", "myom2a", "hey2",
  "col28a1a", "smoc1", "numb",
  "cdh5", "pecam1b", "kdr",
  "mnx1", "LOC125712540", "chata", "elavl3", "rbfox3a"
)
dot_genes <- dot_genes[dot_genes %in% rownames(merged_seurat)]

DotPlot(merged_seurat,
  features = dot_genes,
  group.by = "seurat_clusters"
) +
  RotatedAxis() +
  ggtitle("Canonical cell-type markers across clusters")
```

### Module scores per cell type

`AddModuleScore` computes the average expression of a gene set minus a random control set — positive scores mean the cell expresses that gene set more than expected by chance. We compute one score per cell type plus a PMCA score.

```{r module-score, fig.width=12, fig.height=9, message=FALSE, warning=FALSE, fig.alt="Feature plots on the Harmony UMAP showing module scores for each cell-type gene list and the PMCA stalk score.", fig.cap="Module scores for curated cell-type gene sets projected onto the Harmony UMAP. Warm colors indicate cells enriched for that cell type's markers. The electrocyte score should dominate; the PMCA score highlights stalk candidates; Schwann, endothelial, muscle, and motor neuron scores highlight smaller distinct territories."}
for (ct in names(marker_genes)) {
  if (length(marker_genes[[ct]]) < 3) {
    message(ct, ": fewer than 3 markers found — skipping")
    next
  }
  merged_seurat <- AddModuleScore(
    merged_seurat,
    features = list(marker_genes[[ct]]),
    name     = paste0(ct, "_score"),
    nbin     = 12,
    seed     = 42
  )
  old_col <- paste0(ct, "_score1")
  new_col <- paste0(ct, "_score")
  merged_seurat@meta.data[[new_col]] <- merged_seurat@meta.data[[old_col]]
  merged_seurat@meta.data[[old_col]] <- NULL
}

score_cols <- paste0(names(marker_genes), "_score")
score_cols <- score_cols[score_cols %in% colnames(merged_seurat@meta.data)]

if (length(pmca_genes) >= 2) {
  merged_seurat <- AddModuleScore(
    merged_seurat,
    features = list(pmca_genes),
    name     = "PMCA_score",
    nbin     = 12,
    seed     = 42
  )
  merged_seurat$PMCA_score <- merged_seurat$PMCA_score1
  merged_seurat$PMCA_score1 <- NULL
}

plot_score_cols <- c(
  score_cols,
  if ("PMCA_score" %in% colnames(merged_seurat@meta.data)) "PMCA_score"
)

FeaturePlot(merged_seurat,
  features  = plot_score_cols,
  reduction = "umap.harmony",
  ncol      = 2,
  order     = TRUE
)
```

### Assign cluster labels and refine stalk identity in one pass

For each cluster we pick the cell type with the highest mean module score. Then we sub-classify the electrocyte cluster with the strongest combined PMCA × electrocyte signal as `Stalk_Electrocyte`.

```{r annotate-from-markers, fig.width=9, fig.height=6, message=FALSE, warning=FALSE, fig.alt="UMAP of EO nuclei on the Harmony embedding, clusters labeled by the cell type with the highest module score, with stalk electrocytes refined by PMCA × Electrocyte co-expression.", fig.cap="Annotated EO cell-type atlas. Stalk electrocytes are distinguished from face electrocytes by PMCA module score; Schwann, endothelial, muscle, and motor neuron clusters are identifiable as smaller distinct populations."}
cluster_scores <- merged_seurat@meta.data |>
  group_by(seurat_clusters) |>
  summarise(across(all_of(score_cols), mean, .names = "{.col}"), .groups = "drop")

cluster_labels <- cluster_scores |>
  pivot_longer(-seurat_clusters, names_to = "score_col", values_to = "mean_score") |>
  mutate(cell_type = str_remove(score_col, "_score")) |>
  group_by(seurat_clusters) |>
  slice_max(mean_score, n = 1, with_ties = FALSE) |>
  ungroup() |>
  dplyr::select(seurat_clusters, cell_type, mean_score) |>
  arrange(as.integer(as.character(seurat_clusters)))

label_vec <- setNames(
  cluster_labels$cell_type,
  as.character(cluster_labels$seurat_clusters)
)
merged_seurat$cell_type_go <- unname(label_vec[as.character(merged_seurat$seurat_clusters)])

# Stalk refinement: search ALL clusters by PMCA × electrocyte co-expression score
merged_seurat$cell_type_detail <- merged_seurat$cell_type_go

if (length(pmca_genes) >= 2 && "PMCA_score" %in% colnames(merged_seurat@meta.data)) {
  stalk_scores <- merged_seurat@meta.data |>
    group_by(seurat_clusters) |>
    summarise(
      mean_PMCA = mean(PMCA_score),
      mean_Elec = mean(Electrocyte_score),
      n_cells = n(),
      initial_label = names(sort(table(cell_type_go), decreasing = TRUE))[1],
      .groups = "drop"
    ) |>
    mutate(stalk_score = mean_PMCA * pmax(mean_Elec, 0)) |>
    arrange(desc(stalk_score))

  stalk_cluster <- as.character(stalk_scores$seurat_clusters[1])
  cat("Stalk candidate: cluster", stalk_cluster,
      "(PMCA =", round(stalk_scores$mean_PMCA[1], 3), ")\n")

  pmca_by_cluster <- merged_seurat@meta.data |>
    filter(cell_type_go == "Electrocyte") |>
    group_by(seurat_clusters) |>
    summarise(mean_PMCA = mean(PMCA_score), n_cells = n(), .groups = "drop") |>
    arrange(desc(mean_PMCA))

  stalk_idx <- rownames(merged_seurat@meta.data)[
    as.character(merged_seurat$seurat_clusters) == stalk_cluster
  ]
  merged_seurat$cell_type_detail[stalk_idx] <- "Stalk_Electrocyte"
}

DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "cell_type_detail",
  label     = TRUE,
  repel     = TRUE
) +
  ggtitle("EO cell types (marker-based annotation, Harmony UMAP)") +
  NoLegend()
```

This is a natural checkpoint — the cluster-to-cell-type map is now baked into the object, and the next steps (face sub-classification + pseudobulk) build on it.

```{r save-checkpoint-03, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(merged_seurat,
  file = "data/checkpoints/03_annotated.rds")
```

::: {.callout-important title="Challenge 1: Defend the cell-type calls (15 min)"}

Look at the annotated UMAP and the dot plot above.

1. Which cluster has the strongest electrocyte signature? Is it the stalk or a face cluster?
2. The motor neuron population in EO is often sparse. Did the annotation find one? Inspect the `mnx1` and `chata` FeaturePlot — does the labeled motor neuron cluster express both?
3. Find one cluster you're skeptical of. What evidence would change your mind?

:::: {.callout-tip title="Solution" collapse="true"}

There is no single correct answer — it depends on clustering resolution. General expectations:

- One or more large clusters should receive the **Electrocyte** label, since electrocytes dominate the EO. The cluster refined to **Stalk_Electrocyte** should have high PMCA expression.
- Because electrocytes are myogenic, **Muscle** clusters may appear nearby on the UMAP.
- **Schwann** and **Endothelial** clusters are usually compact and small.
- **Motorneuron** clusters may be sparse — cholinergic terminals don't always survive nuclear isolation. The cluster should co-express *mnx1* (TF) and *chata* (cholinergic identity).

Skepticism is healthy. A "Schwann" cluster with no *col28a1a* expression deserves a second look — likely a misassignment driven by one outlier marker.

::::

:::

::: {.callout-tip title="Stretch break — and a fork in the road"}
This is the end of the cell-type annotation block. The remaining material is denser than what came before it:

- The **face sub-classification** section below (k-means on UMAP centroids, markers, biology check) is conceptually optional — if you're running short on time, **skip directly to "Pseudobulk aggregation"** and come back to this on your own. Anything downstream uses `cell_type_label`, which is already in the saved checkpoint.
- The **DESeq2 + propeller + cross-study scatter** sections are the central statistical payoff and should be hit live.

If you're on pace, take five minutes to stand up, stretch, and refill your water. The rest of the episode runs straight through to differential expression.
:::

## Electrocyte face sub-classification {#sec-face-subclass}

*Optional for fast finishers — skippable if the room is tight on time.* Each electrocyte has two electrochemically distinct membranes: the **posterior (innervated) face**, which generates the action potential, and the **anterior (non-innervated) face**, which generates a counter-potential. The exact channel inventory of each face is not fully resolved in *B. brachyistius*. Our data can begin to address this.

Below we split the non-stalk electrocyte clusters into two candidate face groups using k-means on UMAP centroids, find markers that distinguish them, and judge whether the split is biologically coherent.

```{r elec-face-setup, message=FALSE, warning=FALSE}
if (exists("stalk_cluster") && exists("pmca_by_cluster")) {
  face_clusters <- pmca_by_cluster |>
    filter(as.character(seurat_clusters) != stalk_cluster) |>
    pull(seurat_clusters) |>
    as.character()

  umap_mat <- Embeddings(merged_seurat, "umap.harmony")

  face_centroids <- merged_seurat@meta.data |>
    rownames_to_column("cell") |>
    filter(as.character(seurat_clusters) %in% face_clusters) |>
    left_join(as.data.frame(umap_mat) |> rownames_to_column("cell"), by = "cell") |>
    group_by(seurat_clusters) |>
    summarise(
      U1 = mean(.data[[colnames(umap_mat)[1]]]),
      U2 = mean(.data[[colnames(umap_mat)[2]]]),
      .groups = "drop"
    )

  set.seed(42)
  km <- kmeans(face_centroids[, c("U1", "U2")], centers = 2)
  face_centroids$face_group <- as.character(km$cluster)

  face_A <- face_centroids |> filter(face_group == "1") |> pull(seurat_clusters) |> as.character()
  face_B <- face_centroids |> filter(face_group == "2") |> pull(seurat_clusters) |> as.character()

  cat("Face group A clusters:", paste(face_A, collapse = ", "), "\n")
  cat("Face group B clusters:", paste(face_B, collapse = ", "), "\n")
}
```

```{r elec-face-markers, message=FALSE, warning=FALSE}
if (exists("face_A") && exists("face_B")) {
  elec_cells <- rownames(merged_seurat@meta.data)[
    merged_seurat$cell_type_go == "Electrocyte"
  ]

  face_markers <- FindMarkers(
    merged_seurat,
    ident.1  = face_A,
    ident.2  = face_B,
    group.by = "seurat_clusters",
    cells    = elec_cells,
    verbose  = FALSE
  ) |>
    tibble::rownames_to_column("gene") |>
    left_join(gene_lookup |> dplyr::select(key, gene_name, product),
      by = c("gene" = "key")
    ) |>
    arrange(desc(avg_log2FC))

  head(face_markers |> filter(p_val_adj < 0.05), 15)
}
```

```{r elec-face-categories, message=FALSE, warning=FALSE}
if (exists("face_markers")) {
  face_markers |>
    filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.5, !is.na(product)) |>
    mutate(product = as.character(product),
           category = case_when(
             str_detect(product, regex("potassium|sodium|calcium|chloride|channel|transporter", ignore_case = TRUE)) ~ "Ion channel/transporter",
             str_detect(product, regex("collagen|fibronectin|laminin|cadherin|integrin", ignore_case = TRUE)) ~ "ECM/adhesion",
             str_detect(product, regex("kinase|phosphatase|GTPase|signaling", ignore_case = TRUE)) ~ "Signaling",
             str_detect(product, regex("actin|myosin|tubulin|spectrin|cytoskeletal", ignore_case = TRUE)) ~ "Cytoskeletal",
             str_detect(product, regex("receptor|ligand", ignore_case = TRUE)) ~ "Receptor/ligand",
             TRUE ~ "Other/uncharacterized"
           )) |>
    dplyr::count(category, sort = TRUE)
}
```

::: {.callout-note title="Three reasons electrocyte clusters can split"}

When non-stalk electrocyte clusters form two UMAP-separated groups, the explanation is usually one of three things:

1. **Biological — anterior vs. posterior face identity.** The posterior face is dominated by Nav channels; the anterior face relies on K⁺ and Ca²⁺ channels. Coherent ion-channel families in the top DEGs support this.
2. **Spatial/anatomical — rostral vs. caudal segments.** Some mormyrids show regional EO specialization. Cytoskeletal or ECM differences (not channels) hint at this.
3. **Technical — residual sample-of-origin effects.** Even after Harmony, mild batch can split a cell type. Check sample composition per cluster.

**Rule of thumb:** many between-group DEGs in coherent functional families → biological. Few DEGs or uncharacterized proteins → likely technical.

:::

## Pseudobulk differential expression with DESeq2

This is the central analytical question of the workshop: **does 11-KT change gene expression in the electric organ?**

### Why pseudobulk: the pseudoreplication problem

This dataset contains ~2,900 nuclei from four fish — two 11-KT (BB48, BB49INJ), two vehicle (BB50, BB50INJ). A tempting but **wrong** approach is to test each nucleus as an independent observation: Wilcoxon on ~1,400 11-KT nuclei vs. ~1,400 vehicle nuclei.

The problem: nuclei from the same fish are **not independent**. They share fish-level variation in diet, microbiome, stress history, and countless other factors that have nothing to do with 11-KT. Treating each nucleus as a replicate inflates the degrees of freedom ~700-fold, deflates p-values accordingly, and lets results from one unusual fish look like a reproducible treatment effect.

**Pseudobulk** solves this: we sum raw counts per gene across nuclei within each (cell type × sample), producing one count per gene per cell type per fish. DESeq2 then models those four numbers per cell type with treatment as the contrast.

::: {.callout-warning title="Raw counts only — never SCT residuals or Harmony"}

DESeq2 fits its own size-factor normalization internally. It requires **raw integer count data**:

| Input                                | Appropriate for DE?             |
|--------------------------------------|---------------------------------|
| `AggregateExpression(assay = "RNA")` | ✅ Yes — raw UMI counts         |
| SCT residuals (`scale.data`)         | ❌ No — continuous, not integer |
| Log-normalized counts (`data` slot)  | ❌ No — violates count model    |
| Harmony-corrected embeddings         | ❌ No — not counts at all       |

Use `assay = "RNA"` in every `AggregateExpression` call. SCT and Harmony are for clustering and visualization only.

:::

### Build the pseudobulk matrix

```{r aggregate-expression, message=FALSE, warning=FALSE}
pseudobulk_counts <- AggregateExpression(
  merged_seurat,
  assays        = "RNA",
  group.by      = c("cell_type_go", "orig.ident"),
  return.seurat = FALSE
)$RNA

cat("Pseudobulk matrix dimensions:", dim(pseudobulk_counts), "\n")
print(colnames(pseudobulk_counts))
```

### Sample metadata for the design matrix

Parse the column names (`"CellType_EO-BB48"` etc.) into a metadata data frame with `treatment` as a factor — vehicle as reference so positive log2FC means up in 11-KT.

```{r pseudobulk-metadata, message=FALSE, warning=FALSE}
pb_meta <- data.frame(
  col_id = colnames(pseudobulk_counts),
  stringsAsFactors = FALSE
) |>
  tidyr::separate(col_id,
    into  = c("cell_type", "orig_ident"),
    sep   = "_",
    extra = "merge"
  ) |>
  mutate(
    sample = sub("EO-", "", orig_ident),
    treatment = case_when(
      sample %in% c("BB48", "BB49INJ") ~ "11kt",
      TRUE ~ "vehicle"
    ),
    treatment = factor(treatment, levels = c("vehicle", "11kt"))
  )

rownames(pb_meta) <- colnames(pseudobulk_counts)
pb_meta
```

### Run DESeq2 per cell type

For each cell type with at least 2 samples per condition, build a DESeqDataSet and extract results for the 11-KT vs. vehicle contrast.

```{r filter-clusters, message=FALSE, warning=FALSE}
sample_counts_per_ct <- pb_meta |>
  group_by(cell_type, treatment) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(cell_type) |>
  filter(all(n >= 2)) |>
  pull(cell_type) |>
  unique()

cat("Cell types testable:\n")
print(sample_counts_per_ct)
```

```{r deseq2-loop, message=FALSE, warning=FALSE}
de_results <- list()
dds_objects <- list()

for (ct in sample_counts_per_ct) {
  cols <- which(pb_meta$cell_type == ct)
  mat <- pseudobulk_counts[, cols, drop = FALSE]
  meta_sub <- pb_meta[cols, , drop = FALSE]

  mat <- mat[rowSums(mat) > 0, , drop = FALSE]

  dds <- DESeqDataSetFromMatrix(
    countData = mat,
    colData   = meta_sub,
    design    = ~treatment
  )

  suppressWarnings(dds <- DESeq(dds, quiet = TRUE))

  res <- results(dds,
    contrast = c("treatment", "11kt", "vehicle"),
    alpha    = 0.05
  )

  de_results[[ct]] <- as.data.frame(res) |>
    tibble::rownames_to_column("gene") |>
    mutate(cell_type = ct) |>
    arrange(padj)

  dds_objects[[ct]] <- dds
}

de_combined <- bind_rows(de_results)
cat("Total gene × cell-type tests:", nrow(de_combined), "\n")
```

```{r save-checkpoint-04, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(list(de_results = de_results, dds_objects = dds_objects, pb_meta = pb_meta),
  file = "data/checkpoints/04_pseudobulk_dds.rds")
```

::: {.callout-warning title="Interpreting DESeq2 with n=2 per group"}

With only 2 biological replicates per condition, DESeq2 borrows statistical strength across genes via shrinkage, but uncertainty stays high:

1. **Focus on log2FoldChange magnitude and direction**, not just padj. A gene where both 11-KT fish are higher than both vehicle fish is more credible than one driven by a single outlier.
2. **Padj values will be large** — many genes will not reach 0.05. This is the correct result for an n=2 experiment, not a problem to fix.
3. **A dispersion warning is expected** with n=2 — DESeq2 can't robustly estimate dispersion from two points and borrows from the genome-wide trend.
4. **Treat the output as a ranked list for follow-up**, not a definitive call of DE genes.

:::

### Top DE genes per cell type

```{r de-summary-table, message=FALSE, warning=FALSE}
de_combined |>
  filter(!is.na(padj), padj < 0.1) |>
  group_by(cell_type) |>
  slice_min(padj, n = 5) |>
  ungroup() |>
  left_join(gene_lookup |> dplyr::select(key, product), by = c("gene" = "key")) |>
  dplyr::select(cell_type, gene, product, log2FoldChange, lfcSE, padj)
```

### Per-fish pseudobulk visualization

With n=2 per group, **never plot means ± SE** — show every fish's value as a point. The point of pseudobulk is biological-replicate-level evidence; the visualization should reflect that.

```{r pseudobulk-visualization, fig.width=11, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Dot plots of DESeq2 size-factor-normalized counts for the top 3 DE genes in the Electrocyte cluster, with one dot per fish colored by treatment.", fig.cap="Per-fish pseudobulk expression of top DE genes in the Electrocyte cluster. Each point is one fish (n=2 per condition). Showing individual values rather than means makes the fish-to-fish consistency of each effect transparent."}
ct_vis <- if ("Electrocyte" %in% sample_counts_per_ct) "Electrocyte" else sample_counts_per_ct[1]

dds_vis <- dds_objects[[ct_vis]]
dds_vis <- estimateSizeFactors(dds_vis)
norm_mat <- counts(dds_vis, normalized = TRUE)

top_genes <- de_results[[ct_vis]] |>
  filter(!is.na(padj), padj < 1) |>
  slice_min(padj, n = 3) |>
  pull(gene)

norm_tidy <- norm_mat[top_genes, , drop = FALSE] |>
  as.data.frame() |>
  tibble::rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "col_id", values_to = "norm_count") |>
  left_join(pb_meta |> tibble::rownames_to_column("col_id"), by = "col_id") |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key")
  ) |>
  mutate(
    label = if_else(is.na(product) | product == "",
      gene,
      paste0(gene, "\n(", str_trunc(product, 40), ")")
    )
  )

ggplot(norm_tidy, aes(x = treatment, y = norm_count, color = treatment)) +
  geom_point(size = 4, position = position_jitter(width = 0.08, seed = 42)) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  scale_color_manual(values = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) +
  theme_bw() +
  theme(legend.position = "none", strip.text = element_text(size = 8)) +
  labs(
    title = paste("Pseudobulk normalized counts —", ct_vis, "cluster"),
    y     = "DESeq2-normalized counts",
    x     = "Treatment"
  )
```

::: {.callout-important title="Challenge 2: Pick a gene and defend it (15 min)"}

From the per-fish dot plot, choose **one** top DE gene that you'd present in a lab meeting as a candidate 11-KT-responsive gene.

1. Are both 11-KT fish higher (or both lower) than both vehicle fish? If not, the effect rides on one fish — say so plainly.
2. Look up the `product` description. Does the gene's function make biological sense in the context of EOD lengthening?
3. What would you do next to validate this candidate?

:::: {.callout-tip title="Solution" collapse="true"}

There's no single right answer. The honest reasoning chain:

- **Step 1: Replicate consistency.** If 11-KT fish #1 and #2 sit clearly above (or below) vehicle fish #1 and #2, the effect is reproducible across animals. If one 11-KT fish overlaps a vehicle fish, the result is fragile — one more animal could flip it.
- **Step 2: Biological plausibility.** Genes encoding ion channels (Na⁺, K⁺, Ca²⁺), sarcomeric proteins, or known androgen-pathway components are biologically credible candidates. Lipid-metabolism genes are often systemic androgen responses.
- **Step 3: Validation.** qPCR on independent fish; in situ hybridization to confirm cell-type specificity; published 11-KT response data from related work (you'll do exactly this in the next section against Losilla & Gallant 2025).

::::

:::

## Differential cell-type abundance

Beyond *which genes change*, we can ask whether *cell-type proportions* shift. If 11-KT promotes muscle-to-electrocyte differentiation, the Muscle cluster should shrink and Electrocyte grow.

```{r cell-proportions, message=FALSE, warning=FALSE}
prop_table <- merged_seurat@meta.data |>
  group_by(orig.ident, cell_type_go, treatment, sample) |>
  summarise(n_cells = n(), .groups = "drop") |>
  group_by(orig.ident) |>
  mutate(total = sum(n_cells), prop = n_cells / total) |>
  ungroup()

prop_table |>
  dplyr::select(sample, treatment, cell_type_go, n_cells, total, prop) |>
  arrange(treatment, cell_type_go)
```

```{r prop-scatter, fig.width=9, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Dot plot showing individual fish cell-type proportions for each cluster, colored by treatment.", fig.cap="Individual-fish cell-type proportions. Each point is one fish. With n=2 per condition, visually comparing the four individual values is essential — never plot means or SE."}
ggplot(prop_table,
  aes(x = cell_type_go, y = prop, color = treatment, shape = sample)
) +
  geom_point(size = 3.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Cell-type proportions — individual fish (n = 4)",
    y     = "Proportion of all nuclei",
    x     = "Cell type"
  )
```

### Test for differential abundance (propeller or arcsine-limma fallback)

Both propeller and the arcsine-limma fallback apply an arcsine-square-root transform to proportions before fitting a linear model. With n=2 per group the power is low — report direction and magnitude, not p-values alone.

```{r propeller-or-arcsine, message=FALSE, warning=FALSE}
if (requireNamespace("speckle", quietly = TRUE)) {
  prop_results <- speckle::propeller(
    clusters = merged_seurat$cell_type_go,
    sample   = merged_seurat$orig.ident,
    group    = merged_seurat$treatment
  )
  print(prop_results)
} else {
  message("speckle not available; using arcsine-limma fallback")

  prop_wide <- prop_table |>
    mutate(asin_prop = asin(sqrt(prop))) |>
    dplyr::select(orig.ident, treatment, cell_type_go, asin_prop) |>
    pivot_wider(
      id_cols     = c(orig.ident, treatment),
      names_from  = cell_type_go,
      values_from = asin_prop,
      values_fill = 0
    ) |>
    arrange(treatment)

  design <- model.matrix(~treatment, data = prop_wide)
  ct_cols <- setdiff(names(prop_wide), c("orig.ident", "treatment"))
  fit <- limma::lmFit(t(prop_wide[, ct_cols]), design)
  fit <- limma::eBayes(fit)
  da_res <- limma::topTable(fit, coef = 2, n = Inf)
  print(da_res)
}
```

## Validation: cross-study comparison with Losilla & Gallant 2025

Our pseudobulk result is more trustworthy if it independently recovers the gene-expression changes Losilla & Gallant (2025) found by bulk RNA-seq under 17α-methyltestosterone treatment. We expect: same direction of regulation for shared hits, and a positive correlation in effect sizes.

```{r load-losilla-data, message=FALSE, warning=FALSE}
losilla_bulk <- read_csv(
  "data/ssrnaseq_data/EO/losilla_et_al_data.csv",
  show_col_types = FALSE
) |>
  dplyr::rename(
    gene_bulk  = "Gene Symbol",
    desc_bulk  = "Gene Description",
    logFC_bulk = "logFC",
    FDR_bulk   = "FDR"
  ) |>
  mutate(direction_bulk = if_else(logFC_bulk > 0, "up (androgen)", "down (androgen)"))

cat("Losilla & Gallant 2025 bulk DEGs:", nrow(losilla_bulk), "\n")
```

```{r bulk-snrna-overlap, message=FALSE, warning=FALSE}
ct_check <- if ("Electrocyte" %in% names(de_results)) "Electrocyte" else names(de_results)[1]

overlap_tbl <- de_results[[ct_check]] |>
  inner_join(losilla_bulk, by = c("gene" = "gene_bulk")) |>
  mutate(
    direction_snrna = if_else(log2FoldChange > 0, "up (11-KT)", "down (11-KT)"),
    concordant      = sign(log2FoldChange) == sign(logFC_bulk)
  ) |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product), by = c("gene" = "key"))

n_detected <- nrow(overlap_tbl)
n_concordant <- sum(overlap_tbl$concordant, na.rm = TRUE)

cat(sprintf(
  "Losilla DEGs detected in %s pseudobulk: %d / %d (%.0f%%)\n",
  ct_check, n_detected, nrow(losilla_bulk),
  100 * n_detected / nrow(losilla_bulk)
))
cat(sprintf(
  "Direction concordance: %d / %d (%.0f%%)\n",
  n_concordant, n_detected, 100 * n_concordant / n_detected
))
```

```{r logfc-scatter, fig.width=8, fig.height=6, message=FALSE, warning=FALSE, fig.alt="Scatter plot comparing Losilla & Gallant 2025 bulk logFC (x-axis) against snRNA-seq pseudobulk log2FC (y-axis) for overlapping genes, colored green for concordant direction and pink for discordant.", fig.cap="Bulk vs. single-nucleus fold-change comparison for Losilla DEGs. Each point is one gene detected in both datasets. Green = same direction in both; pink = opposite. A positive trend in the upper-right and lower-left quadrants confirms cross-study reproducibility."}
label_genes <- c(
  "actn1", "actc1c", "myh6", "acta1b", "xirp1",
  "elovl7a", "fhl1a", "atp1b2b",
  "LOC125742960", "LOC125743019", "kcna7", "scn4b"
)

overlap_tbl |>
  mutate(label = if_else(gene %in% label_genes, gene, NA_character_)) |>
  ggplot(aes(x = logFC_bulk, y = log2FoldChange, color = concordant)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(aes(label = label), na.rm = TRUE,
    size = 3, max.overlaps = 20, box.padding = 0.4) +
  scale_color_manual(
    values = c("TRUE" = "#009E73", "FALSE" = "#CC4678"),
    labels = c("TRUE" = "Concordant", "FALSE" = "Discordant"),
    na.value = "grey60"
  ) +
  labs(
    x = "Bulk logFC — Losilla & Gallant 2025\n(T8day vs control, 17αMT)",
    y = "Pseudobulk log2FC — this snRNA-seq\n(11-KT vs vehicle)",
    color = "Direction",
    caption = sprintf("%d genes detected; %d concordant (%.0f%%)",
      n_detected, n_concordant, 100 * n_concordant / n_detected)
  ) +
  theme_bw(base_size = 12)
```

::: {.callout-important title="Challenge 3: Interpreting the cross-study comparison (15 min)"}

Use the scatter plot and concordance numbers to answer:

1. **Direction concordance.** Is the rate substantially above 50% (random chance)? What does that say about whether Harmony + pseudobulk preserved the biological signal?
2. **Effect-size correlation.** Do genes with large bulk logFC also show larger snRNA-seq fold-changes? What does a positive trend mean beyond direction alone?
3. **Hormone comparison.** Losilla used 17α-methyltestosterone (17αMT); we used 11-KT. Both are androgens but differ in receptor selectivity and aromatization potential. Would you predict the same genes responding, a subset, or different ones?

:::: {.callout-tip title="Solution" collapse="true"}

1. **Concordance >70%** is strong evidence that the androgen response was preserved. Because the studies used different androgens, different animals, and different sequencing technologies, high directional agreement provides cross-validation.
2. **A positive trend** means relative effect sizes match across methods — not just direction but magnitude. Genes near the origin in both axes respond weakly in both; the upper-right and lower-left quadrants are robust hits.
3. **Most canonical androgen targets** are expected to respond to both — both act through the androgen receptor. Differences could arise from (a) receptor selectivity — 11-KT is the physiological teleost androgen and may bind with higher affinity than synthetic 17αMT; (b) aromatization — 17αMT cannot be converted to estrogens, whereas 11-KT's aromatization potential differs, possibly activating estrogen-receptor pathways selectively in one study.

::::

:::

::: {.callout-tip title="Key Points"}

1. Cluster markers from `FindAllMarkers` plus NCBI gene-product descriptions are the bridge from cluster number to cell-type identity.
2. Curated marker gene lists from the EO literature, combined with `AddModuleScore`, give defensible cell-type labels in a non-model organism.
3. The stalk sub-classification rides on PMCA paralog expression — Baumann et al. 2025 shows pan-PMCA enrichment on the stalklet membrane.
4. Nuclei from the same fish are not independent replicates; treating them as such is pseudoreplication and inflates false-positive rates.
5. Pseudobulk aggregation (`AggregateExpression` with `assay = "RNA"`) sums raw counts per (cell-type × sample), restoring the n=2-per-group structure for DESeq2.
6. With n=2 per group, report log2FoldChange and direction first, padj second. Plot every fish — never group means.
7. propeller (or arcsine-limma) tests cell-type proportion shifts; with n=2 it is severely underpowered and should be interpreted as a ranked list, not a hypothesis test.
8. Direction concordance >50% with Losilla & Gallant 2025 — and a positive logFC–logFC correlation — confirms that the snRNA-seq pipeline preserved the biological androgen response.

:::
