---
title: "Preprocessing the Electric Organ snRNA-seq Data"
mode: bioinformatics
exercises: 90
optional: false
location: Loeb 156 (Breakroom)       # room shown in the schedule group heading, e.g. Loeb 160
---

::: {.callout-note title="Questions"}

- How do we load a CellRanger output and combine four samples into a single Seurat object?
- Why does SCTransform need to run before PCA, and what does each step contribute?
- When does GEM-well batch correction matter, and how do we check that Harmony actually helped?
- How do we choose a clustering resolution that produces biologically interpretable groups?

:::

::: {.callout-tip title="Objectives"}

- Load CellRanger-derived 10X count matrices for four EO samples and merge them into one Seurat object.
- Attach experimental metadata (sample, fish ID, treatment) before any downstream analysis.
- Apply SCTransform variance-stabilizing normalization and identify highly variable genes.
- Run PCA and read an elbow plot to pick how many PCs to retain.
- Correct for GEM-well batch with Harmony; compare PCA and Harmony UMAPs to confirm the correction.
- Build a kNN graph on the Harmony embedding, cluster with Louvain, and visualize the result by sample, treatment, and cluster.

:::

## Introduction

You watched the QC video and worked through the four-pool worksheet before class. The verdict was unambiguous: **Pool 3 — the electric organ — is the focus of this workshop.** Pool 4 was excluded, Pools 1 and 2 are available for special projects, and the EO library is real, with reasonable Cell Ranger cell calls and intronic fraction that — while lower than ideal — supports analysis.

This episode takes that raw count matrix and walks it through everything you need before you can ask any biological question:

1. Load the four samples and merge them.
2. Stamp every nucleus with its sample, fish, and treatment label.
3. Normalize with SCTransform.
4. Reduce dimensionality with PCA.
5. Correct GEM-well batch with Harmony.
6. Cluster on the Harmony embedding.
7. Visualize and save a checkpoint for Saturday.

By the end of class today you will have an annotated Seurat object with clusters laid out on a UMAP. Tomorrow we use that object to assign cell types and run the 11-KT contrast.

::: {.callout-tip title="Where the data lives"}

The four EO samples live on the workshop server under `data/ssrnaseq_data/EO/`. Each sample (`BB48`, `BB49INJ`, `BB50`, `BB50INJ`) is its own folder of 10X-format files: `barcodes.tsv.gz`, `features.tsv.gz`, `matrix.mtx.gz`. The genome annotation (`genes.gtf.gz`) sits one directory up. You do not need to copy any of this — `Read10X()` reads directly from the folder structure.

:::

## Setup

We use Seurat for everything single-cell, tidyverse for data manipulation, and three specialized packages: `glmGamPoi` (accelerates SCTransform), `presto` (accelerates Wilcoxon for FindMarkers tomorrow), and `harmony` (batch correction).

```{r load-libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(glmGamPoi)
library(presto)
library(harmony)
```

## Load the four EO samples

`Read10X()` reads a 10X-format folder and returns a sparse count matrix; `CreateSeuratObject()` wraps it with metadata structure. We loop over the four samples and keep each as a separate Seurat object for now — the merge step is cleaner that way.

```{r load-data, message=FALSE, warning=FALSE}
tissue <- "EO"
sample_names <- c("BB48", "BB49INJ", "BB50", "BB50INJ")

list_seurat <- list()

for (sample in sample_names) {
  data_dir <- file.path("data/ssrnaseq_data", tissue, sample)

  seurat_data <- Read10X(data.dir = data_dir)
  seurat_obj <- CreateSeuratObject(
    counts = seurat_data,
    min.features = 1,
    project = paste0(tissue, "_", sample)
  )

  list_seurat[[sample]] <- seurat_obj
}
```

::: {.callout-note title="Why we do not filter further"}

Older snRNA-seq tutorials apply heavy manual filtering — minimum UMI thresholds, gene-per-cell cutoffs, mitochondrial-percent caps. We trust Cell Ranger's cell calls in this workshop and skip that step. Two reasons:

- CellRanger's barcode-rank algorithm gave clean cell calls for the EO pool (see the QC worksheet you completed before class — there is a clear knee).
- Manual thresholds on UMI and gene counts encode arbitrary judgment calls. We learn more by reading the Cell Ranger metrics carefully than by re-running a hand-tuned filter in Seurat.

For a production analysis you would also run CellBender for ambient-RNA removal. We skip that here to keep the pipeline self-contained; it is something to revisit before publishing.

:::

## Merge samples and attach metadata

Before any normalization, we merge the four objects into one and stamp each nucleus with the metadata that will travel with it for the rest of the analysis: sample, fish, and treatment.

```{r merge-and-annotate, message=FALSE, warning=FALSE}
# Merge into one object; add.cell.id prefixes each barcode with its sample
merged_seurat <- merge(
  x = list_seurat[[1]],
  y = list_seurat[2:length(list_seurat)],
  add.cell.id = names(list_seurat)
)

# Seurat v5 layers must be joined before downstream analysis
merged_seurat <- JoinLayers(merged_seurat)

# sample: strip the tissue prefix from orig.ident
merged_seurat$sample <- sub("EO_", "", merged_seurat$orig.ident)

# fish: same as sample — each is a different individual
merged_seurat$fish <- merged_seurat$sample

# treatment: assigned by fish ID, NOT by the INJ suffix
# 11-KT treated: BB48, BB49INJ
# Vehicle:       BB50, BB50INJ
treatment_map <- c(
  BB48    = "11kt",
  BB49INJ = "11kt",
  BB50    = "vehicle",
  BB50INJ = "vehicle"
)
merged_seurat$treatment <- unname(treatment_map[merged_seurat$sample])

head(merged_seurat@meta.data)
```

::: {.callout-warning title="The INJ suffix is a library label, not a treatment indicator"}

Two samples are named **BB49INJ** and **BB50INJ**. The "INJ" suffix records a library-preparation detail — it does **not** indicate that those fish received a vehicle injection. Treatment is assigned by fish identity:

| Fish    | Treatment                   |
|---------|-----------------------------|
| BB48    | 11-ketotestosterone (11-KT) |
| BB49INJ | 11-ketotestosterone (11-KT) |
| BB50    | Vehicle (cocoa butter)      |
| BB50INJ | Vehicle (cocoa butter)      |

Using the INJ string to infer treatment would flip the condition label for BB49INJ, producing exactly-wrong results downstream. Always check the experimental metadata, never the sample name.

:::

## Quick visual QC

The QC video covered Cell Ranger metrics in depth. Here we just confirm the per-sample UMI and gene distributions look like what we'd expect — no extra filtering. If one sample looked categorically different from the others (a la Pool 4) we'd stop and reconsider; the EO samples will look reasonable.

```{r qc-violin, fig.width=10, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Violin plots of UMI counts and gene counts per sample.", fig.cap="Distribution of UMI counts (left) and detected genes (right) for each of the four EO samples. Each violin is one sample; points are individual nuclei."}
p_count <- VlnPlot(merged_seurat,
  features = "nCount_RNA",
  group.by = "orig.ident",
  pt.size = 0,
  cols = c("#E69F00", "#56B4E9", "#009E73", "#F0E442")
) +
  theme(legend.position = "none") +
  ggtitle("UMI counts per nucleus")

p_feat <- VlnPlot(merged_seurat,
  features = "nFeature_RNA",
  group.by = "orig.ident",
  pt.size = 0,
  cols = c("#E69F00", "#56B4E9", "#009E73", "#F0E442")
) +
  theme(legend.position = "none") +
  ggtitle("Genes detected per nucleus")

p_count | p_feat
```

::: {.callout-note title="Mitochondrial-percent QC is skipped for this genome"}

In mouse and human data, the percentage of reads from mitochondrial genes (`percent.mt`) flags damaged or lysed cells. Calling `PercentageFeatureSet(merged_seurat, pattern = "^mt-")` here returns approximately zero for every cell — not because the cells are healthy, but because mitochondrial genes in *B. brachyistius* do not use the `^mt-` prefix. Before applying a mitochondrial filter you would need to identify the correct mitochondrial gene names from `features.tsv.gz`. For this workshop we skip that step.

:::

## Normalization with SCTransform

Raw UMI counts vary across nuclei for two reasons: real biology (some cell types transcribe more RNA) and technical noise (nuclei were captured and sequenced at different depths). Normalization removes the technical component so downstream comparisons reflect biology.

We use **SCTransform**, which fits a regularized negative binomial regression per gene. The residuals from those fits are variance-stabilized and depth-corrected; they replace raw counts as the working expression matrix (in a new `SCT` assay).

::: {.callout-note title="Heads up — longest chunk in the lesson"}
SCTransform on four samples takes roughly **3–8 minutes**. The chunk runs with `verbose = FALSE`, so the Console will look quiet — that's expected. Watch the R session indicator (or check `top` in a Terminal) to confirm it's still working before assuming the kernel hung.
:::

```{r sctransform, message=FALSE, warning=FALSE}
merged_seurat <- SCTransform(merged_seurat, verbose = FALSE)
```

::: {.callout-important title="Challenge 1: Inspect the SCT assay (5 min)"}

After SCTransform, the Seurat object has two assays: `RNA` (raw counts) and `SCT` (normalized residuals).

1. How many variable features did SCTransform select? Hint: `length(VariableFeatures(merged_seurat))`.
2. Compare a few values from `merged_seurat[["RNA"]]$counts[1:5, 1:3]` and `merged_seurat[["SCT"]]$scale.data[1:5, 1:3]`. What's different about the SCT values?

:::: {.callout-tip title="Solution" collapse="true"}

```r
length(VariableFeatures(merged_seurat))
# SCTransform selects ~3,000 variable features by default

merged_seurat[["RNA"]]$counts[1:5, 1:3]      # integer UMI counts
merged_seurat[["SCT"]]$scale.data[1:5, 1:3]   # Pearson residuals
```

The SCT scaled values are no longer integers — they are Pearson residuals centered near zero. A gene expressed higher than expected for the nucleus's depth gets a positive residual; lower-than-expected gives negative. These residuals are what PCA operates on.

::::

:::

## Principal component analysis

PCA compresses ~3,000 variable features into a small set of **principal components** — each a weighted combination of genes that captures one axis of variation in the data.

```{r run-pca, message=FALSE, warning=FALSE}
merged_seurat <- RunPCA(merged_seurat, verbose = FALSE, assay = "SCT")
```

The elbow plot ranks PCs by variance explained. We pick enough PCs to be safely past the elbow.

```{r elbow-plot, message=FALSE, warning=FALSE, fig.alt="Elbow plot showing variance explained by each principal component.", fig.cap="Elbow plot for the EO dataset. The variance-explained curve flattens once the elbow is past."}
ElbowPlot(merged_seurat, ndims = 30)
```

::: {.callout-note title="Reading the elbow plot"}

There is no single correct cutoff. Common practice:

- Use enough PCs that the elbow is comfortably included (15–30 for most datasets).
- Err on the side of more — including a few extra noisy PCs rarely hurts, but dropping informative ones can merge biologically distinct clusters.

:::

:::: {.callout-tip title="Workshop choice (predict first!)" collapse="true"}
We use **20 PCs** throughout this episode — comfortably past the elbow and well within the "more is safer than fewer" zone.
::::

We also store an un-integrated UMAP at this point so we can compare it to the Harmony-corrected version a few chunks down. This is purely for the before/after visual check.

```{r run-umap-pca, message=FALSE, warning=FALSE}
merged_seurat <- RunUMAP(
  merged_seurat,
  dims           = 1:20,
  reduction      = "pca",
  reduction.name = "umap.pca",
  verbose        = FALSE
)
```

This is a natural place to **save a checkpoint**. The next steps (Harmony, clustering, UMAP on Harmony) all run quickly, but SCTransform took a few minutes. If you make a mistake later in the episode, restart by loading the checkpoint instead of rerunning SCT.

```{r save-checkpoint-01, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(merged_seurat,
  file = "data/checkpoints/01_merged_sct.rds")
```

## Harmony integration

Each of the four EO samples was captured in its own GEM well, creating potential for **GEM-well batch effects**: small technical differences in ambient RNA, capture efficiency, and lysis conditions. Harmony adjusts the PCA embedding iteratively so that cells align by cell type rather than by capture well.

Importantly, Harmony corrects for *GEM-well identity*, not for *treatment*. Both 11-KT and vehicle fish span both GEM wells, so Harmony cannot and does not remove the treatment signal — only the technical inter-well differences.

```{r harmony-integration, message=FALSE, warning=FALSE}
merged_seurat <- RunHarmony(
  merged_seurat,
  group.by.vars  = "orig.ident",
  reduction      = "pca",
  reduction.save = "harmony",
  verbose        = FALSE
)
```

::: {.callout-warning title="Harmony is a sanity check, not a cure"}

For this dataset, all four samples were processed with the same reagent kit and sequencing run. Technical batch is expected to be small. If the PCA UMAP and the Harmony UMAP look nearly identical, **that is a good result**: it means technical batch was minimal.

A critical constraint: integrated/corrected values are **for clustering and visualization only**. All differential expression tomorrow uses the original raw RNA counts. Passing Harmony embeddings to DESeq2 would violate the count model and give incorrect results.

:::

## Cluster on the Harmony embedding

Seurat clusters cells in two steps:

1. `FindNeighbors` builds a *k*-nearest-neighbor graph in the chosen reduction (here, Harmony).
2. `FindClusters` runs the Louvain algorithm on that graph to find communities — the clusters.

The `resolution` parameter controls how finely the graph is partitioned. We start at `resolution = 0.8` and revisit the choice in a challenge.

```{r harmony-cluster, message=FALSE, warning=FALSE}
merged_seurat <- FindNeighbors(merged_seurat,
  dims      = 1:20,
  reduction = "harmony",
  verbose   = FALSE
)
merged_seurat <- FindClusters(merged_seurat,
  resolution = 0.8,
  verbose    = FALSE
)

# UMAP on the harmony reduction — stored under a named reduction for comparison
merged_seurat <- RunUMAP(
  merged_seurat,
  dims           = 1:20,
  reduction      = "harmony",
  reduction.name = "umap.harmony",
  verbose        = FALSE
)
```

## Before and after: the integration check

This is the diagnostic that tells you whether Harmony helped, hurt, or made no difference. We plot PCA-UMAP and Harmony-UMAP side by side, each colored once by sample and once by treatment.

```{r harmony-umap-comparison, fig.width=14, fig.height=8, message=FALSE, warning=FALSE, fig.alt="A 2-by-2 grid of UMAP plots: PCA and Harmony embeddings, each colored by sample and by treatment.", fig.cap="Un-integrated (PCA) and Harmony-integrated UMAPs, colored by sample (left) and treatment (right). After Harmony correction, sample-specific islands should dissolve while treatment-level separation is preserved. For this low-batch dataset, the two embeddings are often similar — a sign of good sample quality."}
p1 <- DimPlot(merged_seurat,
  reduction = "umap.pca",
  group.by  = "orig.ident",
  alpha     = 0.4
) + ggtitle("PCA — by sample") + theme(legend.position = "right")

p2 <- DimPlot(merged_seurat,
  reduction = "umap.pca",
  group.by  = "treatment",
  alpha     = 0.4,
  cols      = c("11kt" = "#D55E00", "vehicle" = "#0072B2")
) + ggtitle("PCA — by treatment")

p3 <- DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "orig.ident",
  alpha     = 0.4
) + ggtitle("Harmony — by sample") + theme(legend.position = "right")

p4 <- DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "treatment",
  alpha     = 0.4,
  cols      = c("11kt" = "#D55E00", "vehicle" = "#0072B2")
) + ggtitle("Harmony — by treatment")

(p1 | p2) / (p3 | p4)
```

::: {.callout-note title="What to look for"}

**By sample (left column):** sample-specific islands — compact regions occupied by only one or two samples — indicate GEM-well batch. After Harmony (bottom-left) those islands should dissolve.

**By treatment (right column):** 11-KT and vehicle nuclei **should** partially segregate if hormone treatment changes gene expression — that is the biology you want to see. Harmony corrects for GEM-well, not treatment, so treatment-level separation is expected.

If the two rows look nearly identical, that confirms minimal technical batch. Harmony provided assurance, not a dramatic visual change.

:::

## Visualize clusters by sample and treatment

The final view: clusters laid out on the Harmony UMAP, with the same plot colored three ways.

```{r umap-cluster-views, fig.width=15, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Three UMAP panels showing Harmony-corrected nuclei colored by cluster, sample, and treatment.", fig.cap="Harmony UMAP colored by Seurat cluster (left), sample (middle), and treatment (right). Clusters that contain nuclei from all four samples are likely real cell types; clusters dominated by one sample warrant scrutiny."}
p_clu <- DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "seurat_clusters",
  label     = TRUE,
  repel     = TRUE
) + ggtitle("Cluster") + theme(legend.position = "none")

p_smp <- DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "orig.ident",
  alpha     = 0.4
) + ggtitle("Sample")

p_trt <- DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "treatment",
  alpha     = 0.4,
  cols      = c("11kt" = "#D55E00", "vehicle" = "#0072B2")
) + ggtitle("Condition (11-KT vs. vehicle)")

p_clu | p_smp | p_trt
```

::: {.callout-important title="Challenge 2: Resolution sweep (15 min)"}

The `resolution` parameter controls how finely the graph is partitioned. Re-run `FindClusters` at three resolutions and regenerate the UMAP for each.

```r
for (res in c(0.3, 0.8, 1.5)) {
  merged_seurat <- FindClusters(merged_seurat, resolution = res, verbose = FALSE)
  print(
    DimPlot(merged_seurat, reduction = "umap.harmony",
            group.by = "seurat_clusters", label = TRUE) +
      ggtitle(paste("resolution =", res))
  )
}
```

1. How many clusters does each resolution produce?
2. Do the broader clusters at low resolution correspond to merged versions of the fine-grained clusters at high resolution, or are they arranged differently?
3. Which resolution do you expect to be most biologically interpretable for an EO atlas, where you expect a dominant electrocyte population plus a handful of smaller types (muscle, Schwann, endothelial, neurons)?

After exploring, set resolution back to **0.8** so the rest of the pipeline (and tomorrow's annotation) is reproducible.

:::: {.callout-tip title="Solution" collapse="true"}

Low resolution (0.3) gives ~5–8 broad clusters — useful for separating electrocytes from non-electrocytes. High resolution (1.5) can produce 20+ clusters, splitting electrocytes into sub-types (anterior face, posterior face, stalk) and possibly fragmenting the rare populations.

The "right" resolution depends on the question. For the cell-type atlas tomorrow, ~10–14 clusters typically work — enough to resolve stalk vs. face electrocytes without over-splitting the smaller populations. Resolution 0.8 is a defensible starting point.

```r
merged_seurat <- FindClusters(merged_seurat, resolution = 0.8, verbose = FALSE)
```

::::

:::

::: {.callout-important title="Challenge 3 (optional, 10 min): Are any clusters sample-dominated?"}

Use a contingency table to check whether any cluster is more than ~80% from a single sample.

```r
sample_cluster_table <- table(
  merged_seurat$orig.ident,
  merged_seurat$seurat_clusters
)
prop.table(sample_cluster_table, margin = 2) |> round(2)
```

1. Are any clusters more than 80% from a single sample?
2. If yes, where do those clusters sit on the Harmony UMAP — peripheral or central?
3. List two biological and two technical explanations for why a cluster might be sample-dominated.

:::: {.callout-tip title="Solution" collapse="true"}

For the EO dataset most clusters should be reasonably balanced because the dominant cell type (electrocytes) is captured from all four fish. A sample-dominated cluster could indicate:

**Biological:**
- A rare cell type that only happened to be captured in one fish (sampling).
- A real biological state enriched in one treatment group (not an artifact).

**Technical:**
- An ambient RNA or doublet population specific to one capture.
- Residual GEM-well batch that Harmony did not fully correct.

The way to distinguish: tomorrow we look at marker gene expression. Real cell types have coherent markers; technical artifacts usually do not.

::::

:::

## Save the checkpoint for Saturday

Save the preprocessed Seurat object so Saturday's annotation episode can start from a known-good state — and so anyone who falls behind can catch up by loading the checkpoint instead of re-running SCT and Harmony.

```{r save-checkpoint-02, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(merged_seurat,
  file = "data/checkpoints/02_harmony_clustered.rds")
```

::: {.callout-tip title="Key Points"}

1. CellRanger's cell calls are trustworthy for the EO library; we did not apply manual UMI/gene/mitochondrial filtering.
2. Metadata (`sample`, `fish`, `treatment`) must be stamped onto the merged object before any downstream step — the `INJ` suffix is a library label, *not* a treatment indicator.
3. SCTransform normalizes counts using a regularized negative binomial regression; PCA operates on the resulting Pearson residuals.
4. The elbow plot is a judgment call — err on the side of more PCs, and for this dataset 20 is comfortably past the elbow.
5. Harmony corrects GEM-well batch by adjusting the PCA embedding; the side-by-side PCA vs. Harmony UMAP is the diagnostic for whether it helped.
6. Cluster resolution is a parameter, not an answer — try a sweep, then pick the value that gives biologically interpretable groups for your question.
7. Save the checkpoint at `data/checkpoints/02_harmony_clustered.rds` — Saturday's episode loads this object as its starting state.

:::
