---
title: "Cell-Type Annotation and Differential Expression in the Electric Organ"
mode: bioinformatics
exercises: 90
optional: false
location: Loeb 156 (Breakroom)      # room shown in the schedule group heading, e.g. Loeb 160
---

::: {.callout-note title="Questions"}

- How do we turn unlabeled clusters into named cell types using curated marker genes?
- What anatomically distinguishes stalk electrocytes from face electrocytes, and how do we detect that split molecularly?
- Why is it statistically wrong to compare individual nuclei between treatment conditions, and what does pseudobulk fix?
- Does our snRNA-seq pipeline recover the gene-expression signatures of the androgen response previously found by bulk RNA-seq?

:::

::: {.callout-tip title="Objectives"}

- Load the post-Harmony checkpoint produced in Friday's preprocessing episode.
- Find cluster markers with `FindAllMarkers` and join NCBI gene-product descriptions for interpretation.
- Assign cell-type labels (posterior/anterior face, stalk, Schwann, endothelial, satellite, immune) by scoring curated marker panels with `AddModuleScore`.
- Identify the innervated stalk with a falsifiable NMJ-signature prediction rather than by exclusion.
- Aggregate raw counts per (cell type × sample) and run DESeq2 with treatment as the contrast.
- Plot per-fish pseudobulk values for top DE genes and interpret the n=2-per-group result honestly.
- Cross-validate against the Losilla & Gallant 2025 bulk RNA-seq DEG table.

:::

## Introduction

Yesterday you took raw CellRanger output for four EO samples, normalized it, corrected for GEM-well batch with Harmony, and produced a clustered UMAP. The end state — saved as `data/checkpoints/02_harmony_clustered.rds` — is a Seurat object with unlabeled numbered clusters.

Today is the biological payoff. We assign cell-type identities to those clusters, sub-divide the electrocytes into stalk vs. face populations, and run the central statistical test of the experiment: **does 11-KT change gene expression in the electric organ, and if so, in which cells and which genes?**

The episode has four parts:

1. **Cluster markers.** What genes define each cluster, and what do they tell us about cell identity?
2. **Cell-type annotation.** Score curated marker panels and a falsifiable stalk prediction to label every cluster — splitting the electrocytes into anterior and posterior faces and an innervated stalk.
3. **Face validation.** Confirm the posterior/anterior face split with the markers that distinguish them.
4. **Differential expression.** Pseudobulk + DESeq2 contrast 11-KT vs. vehicle within each cell type; cross-validate against Losilla & Gallant 2025.

## Setup and load Friday's checkpoint

The checkpoint is the starting state regardless of whether you ran Friday's full pipeline. If you finished Friday, your in-memory object should already match; loading the checkpoint guarantees a known-good entry point.

```{r load-libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(glmGamPoi)
library(presto)
library(DESeq2)
library(limma)
library(ggrepel)
if (requireNamespace("speckle", quietly = TRUE)) library(speckle)
```

```{r load-checkpoint, message=FALSE, warning=FALSE}
merged_seurat <- readRDS("data/checkpoints/02_harmony_clustered.rds")
merged_seurat
```

## Find cluster markers

::: {.callout-note title="The annotator's job"}
Clustering tells you *which* nuclei group together — it says nothing about *what they are*. For electric-organ tissue there is no off-the-shelf reference atlas to lean on, so we have to earn every label. The method we'll use is the one a good detective uses: treat each cluster as a suspect, form a **falsifiable prediction** about its identity, score the evidence, and revise when the data disagree.
:::

`FindAllMarkers` runs a Wilcoxon test for each cluster against all other cells, returning genes that distinguish each cluster from the rest. Before calling it on SCT-normalized data with multiple samples, we run `PrepSCTFindMarkers()` to re-correct counts to a common depth across samples.

```{r prep-markers, message=FALSE, warning=FALSE}
merged_seurat <- PrepSCTFindMarkers(merged_seurat, verbose = FALSE)
```

```{r find-markers, message=FALSE, warning=FALSE}
all_markers <- FindAllMarkers(
  merged_seurat,
  only.pos = TRUE,
  verbose  = FALSE
)

all_markers <- all_markers |>
  arrange(cluster, desc(avg_log2FC))

head(all_markers, 20)
```

### Annotate with gene-product descriptions

The Seurat row names mix curated gene symbols and `LOC#####` IDs. To make the marker table interpretable, we parse the *B. brachyistius* GTF and join the human-readable product description for each gene.

```{r parse-gtf, message=FALSE, warning=FALSE}
gtf_raw <- read_tsv(
  "data/ssrnaseq_data/genes.gtf.gz",
  comment        = "#",
  col_names      = FALSE,
  col_types      = "ccciicccc",
  show_col_types = FALSE
)

gene_desc <- gtf_raw |>
  filter(str_detect(X9, 'product "')) |>
  mutate(
    gene_id   = str_extract(X9, 'gene_id "([^"]+)"', group = 1),
    gene_name = str_extract(X9, 'gene "([^"]+)"', group = 1),
    product   = str_extract(X9, 'product "([^"]+)"', group = 1)
  ) |>
  dplyr::select(gene_id, gene_name, product) |>
  distinct(gene_id, .keep_all = TRUE)

# Build a combined lookup indexed by either gene_id (LOC ID) OR gene_name (symbol)
gene_lookup <- bind_rows(
  gene_desc |> transmute(key = gene_id, gene_name, product),
  gene_desc |>
    filter(!is.na(gene_name), gene_name != gene_id) |>
    transmute(key = gene_name, gene_name, product)
) |>
  distinct(key, .keep_all = TRUE)
```

```{r annotate-markers, message=FALSE, warning=FALSE}
annotated_markers <- all_markers |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key")
  )

annotated_markers |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 3) |>
  ungroup() |>
  dplyr::select(cluster, gene, gene_name, product, avg_log2FC, p_val_adj) |>
  head(20)
```

::: {.callout-note title="Why so many `LOC#####`?"}

NCBI assigns `LOC` names as provisional symbols when a gene is computationally predicted but lacks an official name. The number is the NCBI Gene ID. The `product` field is where the function lives — "potassium voltage-gated channel...like" is a meaningful clue even when the gene name is `LOC125743019`.

For non-model genomes like *B. brachyistius*, this level of annotation is typical and improves over time as more papers are published.

:::

::: {.callout-tip title="Explore the markers interactively"}
Browse these cluster markers in the [Interactive Marker-Gene Explorer](marker-explorer.html): click a cluster on the UMAP to pull up its sorted, searchable marker table — the same `FindAllMarkers` statistics and product descriptions built above.
:::

## Annotating the clusters: predict, score, revise

Generic GO enrichment performs poorly when one cell type dominates a tissue, and the electric organ is overwhelmingly electrocyte. Instead we annotate with **curated marker panels** — drawn from published work on the mormyrid electric organ and from prior investigation of these exact clusters.

One discipline makes this work in a non-model organism: **a feature's name is the gene symbol only where one exists, and otherwise a provisional `LOC########` ID.** Never type a vertebrate symbol from memory and assume it is present — mine the product descriptions and intersect with the features you actually have. (This is why, below, the canonical "kcna7a" channel appears as `LOC125743019`: there is no `kcna7a` feature in this genome.)

### Reconnaissance: who is an electrocyte?

Before splitting hairs between compartments, the first question is which clusters are even electrocyte-lineage. The principal electrogenic sodium channel **scn4aa** (`LOC125742960`) is the cleanest lineage marker.

::: {.callout-important title="Activity: who is an electrocyte? (6 min)"}
Open the [Marker-Gene Explorer](marker-explorer.html). In the **Gene & Signature** tab, single-gene mode, search **scn4aa** (type `LOC125742960`).

1. Which clusters light up? Eyeball a cutoff around a per-cluster mean of ~2.5 — which clusters clear it?
2. Two clusters sit *just below* the line (c1 ≈ 1.7, c16 ≈ 1.9) and two *just above* (c14 ≈ 2.7, c11 ≈ 3.0). Is 2.5 a hard boundary or a soft one?
3. Now search **dok2**. One cluster lights up brightly. Is it an electrocyte? If not, what might it be?

:::: {.callout-tip title="Solution" collapse="true"}
scn4aa is high across the large electrocyte clusters (both faces and the stalk) plus the posterior-tracking c15. The 2.5 line is **soft** — c11/c14 sit just above, c1/c16 just below — so read it as a guide, not a gate. **c16 is not an electrocyte:** `dok2`, an immune docking protein, peaks there — it is the organ's resident **immune/myeloid** population. And note `scn4aa` is itself only a *lineage* marker (detected in ≈98% of electrocytes): it tells you "electrocyte," not *which* electrocyte cluster.
::::
:::

### Curated marker panels

These panels were worked out from these exact clusters. Each is a small, positive signature we can score per nucleus.

```{r marker-gene-lists, message=FALSE, warning=FALSE}
# Gene-name discipline (see above): PMCA = LOC125738035; Kv1.7-like = LOC125743019;
# IL-8-like = LOC125720678; CCR8-like = LOC125740656; plectin-like = LOC125719088.
marker_genes <- list(
  Posterior_face = c("atp1b2b", "cdh13", "LOC125738035"),          # PMCA corroborating (weakest, ~2x)
  Stalk_NMJ      = c("musk", "chrnb1", "chrna1", "chrne"),         # post-synaptic NMJ apparatus
  Stalk_body     = c("robo2", "slit2", "sptb", "LOC125719088"),    # microstalklet + stalk-proper membrane
  Schwann        = c("col28a1a", "smoc1", "prss12", "numb"),
  Endothelial    = c("cdh5", "pecam1b", "kdr", "flt1"),
  Satellite      = c("pax7b", "notch3", "megf10", "cdh15", "myf5"),# Pax7+ progenitor pool, not muscle
  Immune         = c("dok2", "LOC125720678", "LOC125740656")
)

# Filter to genes actually present in this dataset (gene-name discipline as a safety net)
marker_genes <- lapply(marker_genes, function(g) g[g %in% rownames(merged_seurat)])
print(sapply(marker_genes, length))
```

### Two face programs — and why anterior is defined by absence

Each electrocyte has two electrically distinct membranes. The **posterior (innervated) face** carries a clean, strong program: `atp1b2b` (~70× enriched over the anterior face), `cdh13` (~6–13×), and — more weakly — PMCA (`LOC125738035`, only ~2×). We lead with `atp1b2b`/`cdh13`; PMCA only corroborates.

The **anterior (non-innervated) face** has *no* comparably strong positive marker. It is recognized mainly by the **absence** of the posterior program, with a Kv1.7-like channel (`LOC125743019`) weakly enriched as directional support. That is an honest limitation, not a gap in the analysis — so we score a positive `Posterior_face` panel and read the anterior face as the electrocyte clusters that *lack* it. (The polarity itself — calling the strong program *posterior* — rests partly on contested PMCA protein localization; that two distinct face programs exist is the solid part.)

### Marker dot plot across clusters

A compact diagnostic — 3–4 most informative markers per cell type — projected as a dot plot across clusters. Dot size = fraction of cells expressing; color = mean expression.

```{r marker-dotplot, fig.width=15, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Dot plot showing curated cell-type marker gene expression across Seurat clusters, grouped by electrocyte lineage, posterior and anterior face, stalk NMJ and body, Schwann, endothelial, satellite, and immune programs.", fig.cap="Curated cell-type marker expression across EO clusters. scn4aa/scn4ab mark the electrocyte lineage; atp1b2b/cdh13 (with PMCA, LOC125738035) mark the posterior face while LOC125743019 weakly marks the anterior face; musk/chrnb1/chrna1/utrn mark the stalk NMJ and robo2/slit2/sptb the stalk body; col28a1a/smoc1 mark Schwann, cdh5/pecam1b endothelial, pax7b/notch3/megf10 satellite, and dok2 the immune population."}
dot_genes <- c(
  "LOC125742960", "scn4ab",                  # electrocyte lineage (Na+)
  "atp1b2b", "cdh13", "LOC125738035",        # posterior face (PMCA last = weakest)
  "LOC125743019",                            # anterior face (weak Kv1.7-like)
  "musk", "chrnb1", "chrna1", "utrn",        # stalk — NMJ
  "robo2", "slit2", "sptb",                  # stalk — body
  "col28a1a", "smoc1",                       # Schwann
  "cdh5", "pecam1b",                         # endothelial
  "pax7b", "notch3", "megf10",               # satellite
  "dok2", "LOC125720678"                     # immune
)
dot_genes <- dot_genes[dot_genes %in% rownames(merged_seurat)]

DotPlot(merged_seurat,
  features = dot_genes,
  group.by = "seurat_clusters"
) +
  RotatedAxis() +
  ggtitle("Canonical cell-type markers across clusters")
```

### Module scores per cell type

`AddModuleScore` computes the average expression of a gene set minus a random control set — positive scores mean the cell expresses that gene set more than expected by chance. Pooling several markers this way is also how we rescue **sparse** markers: a low-copy gene like `musk` drops out in many nuclei even when truly on, but the panel as a whole recovers the population no single gene captures. We compute one score per panel.

```{r module-score, fig.width=12, fig.height=9, message=FALSE, warning=FALSE, fig.alt="Feature plots on the Harmony UMAP showing per-panel module scores for the posterior face, stalk NMJ, stalk body, Schwann, endothelial, satellite, and immune programs.", fig.cap="Module scores for the curated panels projected onto the Harmony UMAP. Warm colors indicate nuclei enriched for that panel's markers. The Posterior_face score highlights the innervated-face clusters; the Stalk_NMJ score isolates the subsynaptic stalk cluster; Stalk_body, Schwann, endothelial, satellite, and immune scores mark their smaller distinct territories. The anterior face has no positive panel — it is the electrocyte territory left dark by the Posterior_face score."}
for (ct in names(marker_genes)) {
  if (length(marker_genes[[ct]]) < 3) {
    message(ct, ": fewer than 3 markers found — skipping")
    next
  }
  merged_seurat <- AddModuleScore(
    merged_seurat,
    features = list(marker_genes[[ct]]),
    name     = paste0(ct, "_score"),
    nbin     = 12,
    seed     = 42
  )
  old_col <- paste0(ct, "_score1")
  new_col <- paste0(ct, "_score")
  merged_seurat@meta.data[[new_col]] <- merged_seurat@meta.data[[old_col]]
  merged_seurat@meta.data[[old_col]] <- NULL
}

score_cols <- paste0(names(marker_genes), "_score")
score_cols <- score_cols[score_cols %in% colnames(merged_seurat@meta.data)]

FeaturePlot(merged_seurat,
  features  = score_cols,
  reduction = "umap.harmony",
  ncol      = 2,
  order     = TRUE
)
```

### Predict the stalk before you look

The tempting move is "the stalk is whatever cluster is left over after I've labeled the faces." That is **unfalsifiable** — any leftover cluster would do. Make a positive prediction instead.

::: {.callout-tip title="Define a positive prediction"}
In mormyrids the **stalk is the innervated compartment**, so a stalk nucleus should carry the **post-synaptic neuromuscular apparatus** — acetylcholine-receptor subunits plus MuSK, the master organizer of the synaptic membrane. That is exactly the `Stalk_NMJ` panel. If *no* cluster carries it, we have not found the stalk — an honest result, not a failure.
:::

::: {.callout-important title="Activity: predict the stalk (5 min)"}
In the Explorer's **Signature** mode, load the predefined **Stalk_NMJ** set and read the per-cluster bar chart. Then switch to single-gene mode and search **musk**.

1. Which cluster tops the NMJ signature?
2. Is `musk` specific to that cluster, or spread across many?
3. A reasonable *prior* guess (before the data) was that the stalk would be cluster 12 or 14. Where do 12 and 14 actually rank?

:::: {.callout-tip title="Solution" collapse="true"}
**Cluster 17 tops the NMJ signature decisively, and `musk` is essentially unique to it.** The prior guess (12/14) ranks near the *bottom* — a falsified hypothesis, corrected by the data. c17 is small (44 nuclei, `musk` detected in only ~16), which is exactly why we score the whole panel rather than trust one sparse gene: each NMJ gene drops out in different nuclei, but the module score recovers the population.
::::
:::

We can confirm the ranking directly — average the NMJ module score within each cluster:

```{r nmj-ranking, message=FALSE, warning=FALSE}
nmj_by_cluster <- merged_seurat@meta.data |>
  group_by(seurat_clusters) |>
  summarise(mean_NMJ = mean(Stalk_NMJ_score), n = n(), .groups = "drop") |>
  arrange(desc(mean_NMJ))
head(nmj_by_cluster, 5)

stalk_cluster <- as.character(nmj_by_cluster$seurat_clusters[1])   # -> "17", the NMJ anchor
```

Cluster 17 is the **NMJ / subsynaptic zone** of the stalk — the electrocyte equivalent of the sole-plate nuclei that sit under a muscle synapse and run a different program from the rest of the syncytium. It does not stand alone: on the UMAP it anchors a small **island** (clusters 5, 6, 8, 17) set well apart from every face cluster. Clusters **5 and 6** are two stalk-*body* programs (microstalklets and stalk-proper, marked by `robo2`/`slit2` and `sptb`/plectin); cluster **8** is a ribosomal/OXPHOS-dominated outlier the preprocessing episode flags as likely **contamination**, so we keep it *out* of the stalk.

::: {.callout-note title="Where are the motor neurons?"}
We annotate the *post-synaptic* side of the synapse (c17), but there is no motor-neuron cluster — and that is exactly right. A motor neuron's soma, and therefore its nucleus, sits in the hindbrain/spinal cord; only its distal axon terminal reaches the organ. The cell body simply isn't in the EO sample, so its absence **corroborates** the c17 = NMJ call rather than undercutting it. What we *do* capture of the neuron's escort is the resident Schwann cells (c4/c10).
:::

### Commit the annotation

Having scored every panel and tested the stalk prediction, we can write down the cluster-to-cell-type map. This is the synthesis step: the labels are an explicit, evidence-backed decision, not the output of a fragile one-line `argmax`.

```{r annotate-from-markers, fig.width=9, fig.height=6, message=FALSE, warning=FALSE, fig.alt="UMAP of EO nuclei on the Harmony embedding, clusters labeled by evidence-based cell-type identity: posterior face, anterior face, stalk (NMJ and body zones), Schwann, endothelial, satellite, immune, and a quarantined low-confidence group.", fig.cap="Annotated EO cell-type atlas. Faces are split posterior/anterior; the stalk island (5/6/17) is shown with its NMJ vs body zones; Schwann, endothelial, satellite, and immune populations are smaller distinct territories. Clusters 1, 8, and 11 (progenitor-like, ribosomal contamination, and unresolved) are quarantined as low-confidence."}
# Evidence-derived cluster identities.
# Faces split posterior/anterior; the stalk island (5/6/17) is merged to one label for
# downstream pseudobulk. c8 (ribosomal contamination), c1 (progenitor-like, below the
# scn4aa cutoff) and c11 (unresolved electrocyte state) are quarantined as "Other".
cluster_identity <- c(
  "0"  = "Posterior_face", "3"  = "Posterior_face", "15" = "Posterior_face",
  "2"  = "Anterior_face",  "7"  = "Anterior_face",  "12" = "Anterior_face", "14" = "Anterior_face",
  "5"  = "Stalk",          "6"  = "Stalk",          "17" = "Stalk",
  "4"  = "Schwann",        "10" = "Schwann",
  "9"  = "Endothelial",
  "13" = "Satellite",
  "16" = "Immune",
  "1"  = "Other", "8" = "Other", "11" = "Other"
)
merged_seurat$cell_type_go <- unname(
  cluster_identity[as.character(merged_seurat$seurat_clusters)]
)

# Finer view for the atlas plot: split the stalk into its NMJ vs body zones
cluster_detail <- cluster_identity
cluster_detail[c("5", "6")] <- "Stalk_body"
cluster_detail["17"]        <- "Stalk_NMJ"
merged_seurat$cell_type_detail <- unname(
  cluster_detail[as.character(merged_seurat$seurat_clusters)]
)

print(table(merged_seurat$cell_type_go))

DimPlot(merged_seurat,
  reduction = "umap.harmony",
  group.by  = "cell_type_detail",
  label     = TRUE,
  repel     = TRUE
) +
  ggtitle("EO cell types (evidence-based annotation, Harmony UMAP)") +
  NoLegend()
```

This is a natural checkpoint — the cluster-to-cell-type map is now baked into the object, and the next steps (face sub-classification + pseudobulk) build on it.

```{r save-checkpoint-03, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(merged_seurat,
  file = "data/checkpoints/03_annotated.rds")
```

::: {.callout-important title="Challenge 1: Stress-test the cell-type calls (15 min)"}

Use the dot plot, the module-score UMAPs, and the Explorer to push back on the map we just committed.

1. **The stalk call.** We leaned heavily on cluster 17 despite its small size. Search `musk` in the Explorer: how specific is it to c17 (peak vs. everywhere else)? Given only ~16 of its 44 nuclei are `musk`⁺, why is the module score still trustworthy?
2. **The anterior face.** Posterior face has `atp1b2b`/`cdh13`; what positive marker cleanly defines the *anterior* face? Look at the `Posterior_face` module-score UMAP — how is the anterior face actually recognized?
3. **A quarantined cluster.** We dumped clusters 1, 8, and 11 into "Other." Pick one, inspect its top markers in the Explorer (Tab 2), and decide: is it a real cell type we mislabeled, contamination, or genuinely unresolved? What evidence would change your mind?

:::: {.callout-tip title="Solution" collapse="true"}

- **Stalk (c17).** `musk` peaks at c17 and is near-zero elsewhere, so its *specificity* (the gap between pct-in and pct-out) is high even though its *sensitivity* is low — a classic sparse, low-copy marker. Pooling the whole NMJ panel into a module score recovers the population that any single gene misses through dropout.
- **Anterior face.** There is *no* strong positive marker — `LOC125743019` (Kv1.7-like) is only weakly enriched. The anterior face is the electrocyte territory that stays **dark on the `Posterior_face` score**: it's defined by the *absence* of the posterior program, which is an honest limitation of the data, not a labeling mistake.
- **Other (1/8/11).** c8 is the ribosomal/OXPHOS outlier (~9% ribosomal, ~half from one fish) — the classic contamination signature, not a cell type. c1 sits below the `scn4aa` lineage cutoff with progenitor-like markers. c11 is electrocyte-lineage but has no coherent positive program — genuinely unresolved. Quarantining ambiguous clusters rather than force-labeling them is good practice: it keeps the downstream differential-expression groups clean.

::::

:::

::: {.callout-tip title="Stretch break — and a fork in the road"}
This is the end of the cell-type annotation block. The remaining material is denser than what came before it:

- The **face sub-classification** section below (markers that distinguish posterior from anterior, biology check) is conceptually optional — if you're running short on time, **skip directly to "Pseudobulk aggregation"** and come back to this on your own. Anything downstream uses `cell_type_go`, which is already in the saved checkpoint.
- The **DESeq2 + propeller + cross-study scatter** sections are the central statistical payoff and should be hit live.

If you're on pace, take five minutes to stand up, stretch, and refill your water. The rest of the episode runs straight through to differential expression.
:::

## Electrocyte face sub-classification {#sec-face-subclass}

*Optional for fast finishers — skippable if the room is tight on time.* We already split the electrocytes into a **posterior (innervated) face** and an **anterior (non-innervated) face** during annotation. Here we *validate* that split. The posterior face generates the action potential (Nav-dominated); the anterior face generates a counter-potential. The exact channel inventory of each face is not fully resolved in *B. brachyistius* — our data can begin to address it.

Below we find the genes that distinguish the two labeled face groups and judge whether they form a biologically coherent program (a positive `avg_log2FC` means up in the posterior face).

```{r elec-face-markers, message=FALSE, warning=FALSE}
if (all(c("Posterior_face", "Anterior_face") %in% merged_seurat$cell_type_go)) {
  face_markers <- FindMarkers(
    merged_seurat,
    ident.1  = "Posterior_face",
    ident.2  = "Anterior_face",
    group.by = "cell_type_go",
    verbose  = FALSE
  ) |>
    tibble::rownames_to_column("gene") |>
    left_join(gene_lookup |> dplyr::select(key, gene_name, product),
      by = c("gene" = "key")
    ) |>
    arrange(desc(avg_log2FC))

  head(face_markers |> filter(p_val_adj < 0.05), 15)
}
```

```{r elec-face-categories, message=FALSE, warning=FALSE}
if (exists("face_markers")) {
  face_markers |>
    filter(p_val_adj < 0.05, abs(avg_log2FC) > 0.5, !is.na(product)) |>
    mutate(product = as.character(product),
           category = case_when(
             str_detect(product, regex("potassium|sodium|calcium|chloride|channel|transporter", ignore_case = TRUE)) ~ "Ion channel/transporter",
             str_detect(product, regex("collagen|fibronectin|laminin|cadherin|integrin", ignore_case = TRUE)) ~ "ECM/adhesion",
             str_detect(product, regex("kinase|phosphatase|GTPase|signaling", ignore_case = TRUE)) ~ "Signaling",
             str_detect(product, regex("actin|myosin|tubulin|spectrin|cytoskeletal", ignore_case = TRUE)) ~ "Cytoskeletal",
             str_detect(product, regex("receptor|ligand", ignore_case = TRUE)) ~ "Receptor/ligand",
             TRUE ~ "Other/uncharacterized"
           )) |>
    dplyr::count(category, sort = TRUE)
}
```

::: {.callout-note title="Three reasons electrocyte clusters can split"}

When non-stalk electrocyte clusters form two UMAP-separated groups, the explanation is usually one of three things:

1. **Biological — anterior vs. posterior face identity.** The posterior face is dominated by Nav channels; the anterior face relies on K⁺ and Ca²⁺ channels. Coherent ion-channel families in the top DEGs support this.
2. **Spatial/anatomical — rostral vs. caudal segments.** Some mormyrids show regional EO specialization. Cytoskeletal or ECM differences (not channels) hint at this.
3. **Technical — residual sample-of-origin effects.** Even after Harmony, mild batch can split a cell type. Check sample composition per cluster.

**Rule of thumb:** many between-group DEGs in coherent functional families → biological. Few DEGs or uncharacterized proteins → likely technical.

:::

## Pseudobulk differential expression with DESeq2

This is the central analytical question of the workshop: **does 11-KT change gene expression in the electric organ?**

### Why pseudobulk: the pseudoreplication problem

This dataset contains ~2,900 nuclei from four fish — two 11-KT (BB48, BB49INJ), two vehicle (BB50, BB50INJ). A tempting but **wrong** approach is to test each nucleus as an independent observation: Wilcoxon on ~1,400 11-KT nuclei vs. ~1,400 vehicle nuclei.

The problem: nuclei from the same fish are **not independent**. They share fish-level variation in diet, microbiome, stress history, and countless other factors that have nothing to do with 11-KT. Treating each nucleus as a replicate inflates the degrees of freedom ~700-fold, deflates p-values accordingly, and lets results from one unusual fish look like a reproducible treatment effect.

**Pseudobulk** solves this: we sum raw counts per gene across nuclei within each (cell type × sample), producing one count per gene per cell type per fish. DESeq2 then models those four numbers per cell type with treatment as the contrast.

::: {.callout-warning title="Raw counts only — never SCT residuals or Harmony"}

DESeq2 fits its own size-factor normalization internally. It requires **raw integer count data**:

| Input                                | Appropriate for DE?             |
|--------------------------------------|---------------------------------|
| `AggregateExpression(assay = "RNA")` | ✅ Yes — raw UMI counts         |
| SCT residuals (`scale.data`)         | ❌ No — continuous, not integer |
| Log-normalized counts (`data` slot)  | ❌ No — violates count model    |
| Harmony-corrected embeddings         | ❌ No — not counts at all       |

Use `assay = "RNA"` in every `AggregateExpression` call. SCT and Harmony are for clustering and visualization only.

:::

### Build the pseudobulk matrix

```{r aggregate-expression, message=FALSE, warning=FALSE}
pseudobulk_counts <- AggregateExpression(
  merged_seurat,
  assays        = "RNA",
  group.by      = c("cell_type_go", "orig.ident"),
  return.seurat = FALSE
)$RNA

cat("Pseudobulk matrix dimensions:", dim(pseudobulk_counts), "\n")
print(colnames(pseudobulk_counts))
```

### Sample metadata for the design matrix

Parse the column names (`"CellType_EO-BB48"` etc.) into a metadata data frame with `treatment` as a factor — vehicle as reference so positive log2FC means up in 11-KT.

```{r pseudobulk-metadata, message=FALSE, warning=FALSE}
pb_meta <- data.frame(
  col_id = colnames(pseudobulk_counts),
  stringsAsFactors = FALSE
) |>
  tidyr::separate(col_id,
    into  = c("cell_type", "orig_ident"),
    sep   = "_",
    extra = "merge"
  ) |>
  mutate(
    # AggregateExpression sanitizes "_" in group labels to "-"; restore underscores so
    # cell_type matches the annotation-side labels (Posterior_face, Anterior_face, ...).
    cell_type = gsub("-", "_", cell_type),
    sample = sub("EO-", "", orig_ident),
    treatment = case_when(
      sample %in% c("BB48", "BB49INJ") ~ "11kt",
      TRUE ~ "vehicle"
    ),
    treatment = factor(treatment, levels = c("vehicle", "11kt"))
  )

rownames(pb_meta) <- colnames(pseudobulk_counts)
pb_meta
```

### Run DESeq2 per cell type

For each cell type with at least 2 samples per condition, build a DESeqDataSet and extract results for the 11-KT vs. vehicle contrast.

```{r filter-clusters, message=FALSE, warning=FALSE}
sample_counts_per_ct <- pb_meta |>
  group_by(cell_type, treatment) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(cell_type) |>
  filter(all(n >= 2)) |>
  pull(cell_type) |>
  unique() |>
  setdiff("Other")   # "Other" pools contamination/ambiguous clusters — not a biological group

cat("Cell types testable:\n")
print(sample_counts_per_ct)
```

```{r deseq2-loop, message=FALSE, warning=FALSE}
de_results <- list()
dds_objects <- list()

for (ct in sample_counts_per_ct) {
  cols <- which(pb_meta$cell_type == ct)
  mat <- pseudobulk_counts[, cols, drop = FALSE]
  meta_sub <- pb_meta[cols, , drop = FALSE]

  mat <- mat[rowSums(mat) > 0, , drop = FALSE]

  dds <- DESeqDataSetFromMatrix(
    countData = mat,
    colData   = meta_sub,
    design    = ~treatment
  )

  suppressWarnings(dds <- DESeq(dds, quiet = TRUE))

  res <- results(dds,
    contrast = c("treatment", "11kt", "vehicle"),
    alpha    = 0.05
  )

  de_results[[ct]] <- as.data.frame(res) |>
    tibble::rownames_to_column("gene") |>
    mutate(cell_type = ct) |>
    arrange(padj)

  dds_objects[[ct]] <- dds
}

de_combined <- bind_rows(de_results)
cat("Total gene × cell-type tests:", nrow(de_combined), "\n")
```

```{r save-checkpoint-04, message=FALSE, warning=FALSE, eval=FALSE}
saveRDS(list(de_results = de_results, dds_objects = dds_objects, pb_meta = pb_meta),
  file = "data/checkpoints/04_pseudobulk_dds.rds")
```

::: {.callout-warning title="Interpreting DESeq2 with n=2 per group"}

With only 2 biological replicates per condition, DESeq2 borrows statistical strength across genes via shrinkage, but uncertainty stays high:

1. **Focus on log2FoldChange magnitude and direction**, not just padj. A gene where both 11-KT fish are higher than both vehicle fish is more credible than one driven by a single outlier.
2. **Padj values will be large** — many genes will not reach 0.05. This is the correct result for an n=2 experiment, not a problem to fix.
3. **A dispersion warning is expected** with n=2 — DESeq2 can't robustly estimate dispersion from two points and borrows from the genome-wide trend.
4. **Treat the output as a ranked list for follow-up**, not a definitive call of DE genes.

:::

### Top DE genes per cell type

```{r de-summary-table, message=FALSE, warning=FALSE}
de_combined |>
  filter(!is.na(padj), padj < 0.1) |>
  group_by(cell_type) |>
  slice_min(padj, n = 5) |>
  ungroup() |>
  left_join(gene_lookup |> dplyr::select(key, product), by = c("gene" = "key")) |>
  dplyr::select(cell_type, gene, product, log2FoldChange, lfcSE, padj)
```

### Per-fish pseudobulk visualization

With n=2 per group, **never plot means ± SE** — show every fish's value as a point. The point of pseudobulk is biological-replicate-level evidence; the visualization should reflect that.

```{r pseudobulk-visualization, fig.width=11, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Dot plots of DESeq2 size-factor-normalized counts for the top 3 DE genes in the visualized cell type, with one dot per fish colored by treatment.", fig.cap="Per-fish pseudobulk expression of top DE genes in the visualized cell type (ct_vis). Each point is one fish (n=2 per condition). Showing individual values rather than means makes the fish-to-fish consistency of each effect transparent."}
ct_vis <- if ("Posterior_face" %in% sample_counts_per_ct) "Posterior_face" else sample_counts_per_ct[1]

dds_vis <- dds_objects[[ct_vis]]
dds_vis <- estimateSizeFactors(dds_vis)
norm_mat <- counts(dds_vis, normalized = TRUE)

top_genes <- de_results[[ct_vis]] |>
  filter(!is.na(padj), padj < 1) |>
  slice_min(padj, n = 3) |>
  pull(gene)

norm_tidy <- norm_mat[top_genes, , drop = FALSE] |>
  as.data.frame() |>
  tibble::rownames_to_column("gene") |>
  pivot_longer(-gene, names_to = "col_id", values_to = "norm_count") |>
  left_join(pb_meta |> tibble::rownames_to_column("col_id"), by = "col_id") |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product),
    by = c("gene" = "key")
  ) |>
  mutate(
    label = if_else(is.na(product) | product == "",
      gene,
      paste0(gene, "\n(", str_trunc(product, 40), ")")
    )
  )

ggplot(norm_tidy, aes(x = treatment, y = norm_count, color = treatment)) +
  geom_point(size = 4, position = position_jitter(width = 0.08, seed = 42)) +
  facet_wrap(~label, scales = "free_y", nrow = 1) +
  scale_color_manual(values = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) +
  theme_bw() +
  theme(legend.position = "none", strip.text = element_text(size = 8)) +
  labs(
    title = paste("Pseudobulk normalized counts —", ct_vis, "cluster"),
    y     = "DESeq2-normalized counts",
    x     = "Treatment"
  )
```

::: {.callout-important title="Challenge 2: Pick a gene and defend it (15 min)"}

From the per-fish dot plot, choose **one** top DE gene that you'd present in a lab meeting as a candidate 11-KT-responsive gene.

1. Are both 11-KT fish higher (or both lower) than both vehicle fish? If not, the effect rides on one fish — say so plainly.
2. Look up the `product` description. Does the gene's function make biological sense in the context of EOD lengthening?
3. What would you do next to validate this candidate?

:::: {.callout-tip title="Solution" collapse="true"}

There's no single right answer. The honest reasoning chain:

- **Step 1: Replicate consistency.** If 11-KT fish #1 and #2 sit clearly above (or below) vehicle fish #1 and #2, the effect is reproducible across animals. If one 11-KT fish overlaps a vehicle fish, the result is fragile — one more animal could flip it.
- **Step 2: Biological plausibility.** Genes encoding ion channels (Na⁺, K⁺, Ca²⁺), sarcomeric proteins, or known androgen-pathway components are biologically credible candidates. Lipid-metabolism genes are often systemic androgen responses.
- **Step 3: Validation.** qPCR on independent fish; in situ hybridization to confirm cell-type specificity; published 11-KT response data from related work (you'll do exactly this in the next section against Losilla & Gallant 2025).

::::

:::

## Differential cell-type abundance

Beyond *which genes change*, we can ask whether *cell-type proportions* shift. If 11-KT drives the satellite pool to differentiate into electrocytes, the Satellite fraction should shrink and the face populations grow.

```{r cell-proportions, message=FALSE, warning=FALSE}
prop_table <- merged_seurat@meta.data |>
  group_by(orig.ident, cell_type_go, treatment, sample) |>
  summarise(n_cells = n(), .groups = "drop") |>
  group_by(orig.ident) |>
  mutate(total = sum(n_cells), prop = n_cells / total) |>
  ungroup()

prop_table |>
  dplyr::select(sample, treatment, cell_type_go, n_cells, total, prop) |>
  arrange(treatment, cell_type_go)
```

```{r prop-scatter, fig.width=9, fig.height=5, message=FALSE, warning=FALSE, fig.alt="Dot plot showing individual fish cell-type proportions for each cluster, colored by treatment.", fig.cap="Individual-fish cell-type proportions. Each point is one fish. With n=2 per condition, visually comparing the four individual values is essential — never plot means or SE."}
ggplot(prop_table,
  aes(x = cell_type_go, y = prop, color = treatment, shape = sample)
) +
  geom_point(size = 3.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("11kt" = "#D55E00", "vehicle" = "#0072B2")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Cell-type proportions — individual fish (n = 4)",
    y     = "Proportion of all nuclei",
    x     = "Cell type"
  )
```

### Test for differential abundance (propeller or arcsine-limma fallback)

Both propeller and the arcsine-limma fallback apply an arcsine-square-root transform to proportions before fitting a linear model. With n=2 per group the power is low — report direction and magnitude, not p-values alone.

```{r propeller-or-arcsine, message=FALSE, warning=FALSE}
if (requireNamespace("speckle", quietly = TRUE)) {
  prop_results <- speckle::propeller(
    clusters = merged_seurat$cell_type_go,
    sample   = merged_seurat$orig.ident,
    group    = merged_seurat$treatment
  )
  print(prop_results)
} else {
  message("speckle not available; using arcsine-limma fallback")

  prop_wide <- prop_table |>
    mutate(asin_prop = asin(sqrt(prop))) |>
    dplyr::select(orig.ident, treatment, cell_type_go, asin_prop) |>
    pivot_wider(
      id_cols     = c(orig.ident, treatment),
      names_from  = cell_type_go,
      values_from = asin_prop,
      values_fill = 0
    ) |>
    arrange(treatment)

  design <- model.matrix(~treatment, data = prop_wide)
  ct_cols <- setdiff(names(prop_wide), c("orig.ident", "treatment"))
  fit <- limma::lmFit(t(prop_wide[, ct_cols]), design)
  fit <- limma::eBayes(fit)
  da_res <- limma::topTable(fit, coef = 2, n = Inf)
  print(da_res)
}
```

## Validation: cross-study comparison with Losilla & Gallant 2025

Our pseudobulk result is more trustworthy if it independently recovers the gene-expression changes Losilla & Gallant (2025) found by bulk RNA-seq under 17α-methyltestosterone treatment. We expect: same direction of regulation for shared hits, and a positive correlation in effect sizes.

```{r load-losilla-data, message=FALSE, warning=FALSE}
losilla_bulk <- read_csv(
  "data/ssrnaseq_data/EO/losilla_et_al_data.csv",
  show_col_types = FALSE
) |>
  dplyr::rename(
    gene_bulk  = "Gene Symbol",
    desc_bulk  = "Gene Description",
    logFC_bulk = "logFC",
    FDR_bulk   = "FDR"
  ) |>
  mutate(direction_bulk = if_else(logFC_bulk > 0, "up (androgen)", "down (androgen)"))

cat("Losilla & Gallant 2025 bulk DEGs:", nrow(losilla_bulk), "\n")
```

```{r bulk-snrna-overlap, message=FALSE, warning=FALSE}
ct_check <- if ("Posterior_face" %in% names(de_results)) "Posterior_face" else names(de_results)[1]

overlap_tbl <- de_results[[ct_check]] |>
  inner_join(losilla_bulk, by = c("gene" = "gene_bulk")) |>
  mutate(
    direction_snrna = if_else(log2FoldChange > 0, "up (11-KT)", "down (11-KT)"),
    concordant      = sign(log2FoldChange) == sign(logFC_bulk)
  ) |>
  left_join(gene_lookup |> dplyr::select(key, gene_name, product), by = c("gene" = "key"))

n_detected <- nrow(overlap_tbl)
n_concordant <- sum(overlap_tbl$concordant, na.rm = TRUE)

cat(sprintf(
  "Losilla DEGs detected in %s pseudobulk: %d / %d (%.0f%%)\n",
  ct_check, n_detected, nrow(losilla_bulk),
  100 * n_detected / nrow(losilla_bulk)
))
cat(sprintf(
  "Direction concordance: %d / %d (%.0f%%)\n",
  n_concordant, n_detected, 100 * n_concordant / n_detected
))
```

```{r logfc-scatter, fig.width=8, fig.height=6, message=FALSE, warning=FALSE, fig.alt="Scatter plot comparing Losilla & Gallant 2025 bulk logFC (x-axis) against snRNA-seq pseudobulk log2FC (y-axis) for overlapping genes, colored green for concordant direction and pink for discordant.", fig.cap="Bulk vs. single-nucleus fold-change comparison for Losilla DEGs. Each point is one gene detected in both datasets. Green = same direction in both; pink = opposite. A positive trend in the upper-right and lower-left quadrants confirms cross-study reproducibility."}
label_genes <- c(
  "actn1", "actc1c", "myh6", "acta1b", "xirp1",
  "elovl7a", "fhl1a", "atp1b2b",
  "LOC125742960", "LOC125743019", "kcna7", "scn4b"
)

overlap_tbl |>
  mutate(label = if_else(gene %in% label_genes, gene, NA_character_)) |>
  ggplot(aes(x = logFC_bulk, y = log2FoldChange, color = concordant)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  ggrepel::geom_text_repel(aes(label = label), na.rm = TRUE,
    size = 3, max.overlaps = 20, box.padding = 0.4) +
  scale_color_manual(
    values = c("TRUE" = "#009E73", "FALSE" = "#CC4678"),
    labels = c("TRUE" = "Concordant", "FALSE" = "Discordant"),
    na.value = "grey60"
  ) +
  labs(
    x = "Bulk logFC — Losilla & Gallant 2025\n(T8day vs control, 17αMT)",
    y = "Pseudobulk log2FC — this snRNA-seq\n(11-KT vs vehicle)",
    color = "Direction",
    caption = sprintf("%d genes detected; %d concordant (%.0f%%)",
      n_detected, n_concordant, 100 * n_concordant / n_detected)
  ) +
  theme_bw(base_size = 12)
```

::: {.callout-important title="Challenge 3: Interpreting the cross-study comparison (15 min)"}

Use the scatter plot and concordance numbers to answer:

1. **Direction concordance.** Is the rate substantially above 50% (random chance)? What does that say about whether Harmony + pseudobulk preserved the biological signal?
2. **Effect-size correlation.** Do genes with large bulk logFC also show larger snRNA-seq fold-changes? What does a positive trend mean beyond direction alone?
3. **Hormone comparison.** Losilla used 17α-methyltestosterone (17αMT); we used 11-KT. Both are androgens but differ in receptor selectivity and aromatization potential. Would you predict the same genes responding, a subset, or different ones?

:::: {.callout-tip title="Solution" collapse="true"}

1. **Concordance >70%** is strong evidence that the androgen response was preserved. Because the studies used different androgens, different animals, and different sequencing technologies, high directional agreement provides cross-validation.
2. **A positive trend** means relative effect sizes match across methods — not just direction but magnitude. Genes near the origin in both axes respond weakly in both; the upper-right and lower-left quadrants are robust hits.
3. **Most canonical androgen targets** are expected to respond to both — both act through the androgen receptor. Differences could arise from (a) receptor selectivity — 11-KT is the physiological teleost androgen and may bind with higher affinity than synthetic 17αMT; (b) aromatization — 17αMT cannot be converted to estrogens, whereas 11-KT's aromatization potential differs, possibly activating estrogen-receptor pathways selectively in one study.

::::

:::

::: {.callout-tip title="Key Points"}

1. Cluster markers from `FindAllMarkers` plus NCBI gene-product descriptions are the bridge from cluster number to cell-type identity.
2. Curated marker gene lists from the EO literature, combined with `AddModuleScore`, give defensible cell-type labels in a non-model organism.
3. A *falsifiable prediction* — the stalk should carry the post-synaptic NMJ apparatus (`musk`/AChR) — beats labeling by exclusion: it points cleanly to the subsynaptic cluster (17) and is corroborated by the absent motor-neuron soma.
4. Nuclei from the same fish are not independent replicates; treating them as such is pseudoreplication and inflates false-positive rates.
5. Pseudobulk aggregation (`AggregateExpression` with `assay = "RNA"`) sums raw counts per (cell-type × sample), restoring the n=2-per-group structure for DESeq2.
6. With n=2 per group, report log2FoldChange and direction first, padj second. Plot every fish — never group means.
7. propeller (or arcsine-limma) tests cell-type proportion shifts; with n=2 it is severely underpowered and should be interpreted as a ranked list, not a hypothesis test.
8. Direction concordance >50% with Losilla & Gallant 2025 — and a positive logFC–logFC correlation — confirms that the snRNA-seq pipeline preserved the biological androgen response.

:::
